The cow udder is a potential coinfection site for influenza A viruses

Version 1

Abstract

The incursion of high pathogenicity avian influenza A virus (IAV) into US dairy cows is unprecedented in the era of molecular diagnosis and pathogen sequencing. This raises questions over the likelihood of further outbreaks and whether dairy cattle could be a “mixing vessel” for novel strains of IAV. Using a panel of BSL2-safe reassortant viruses representing clade 2.3.4.4b H5 epizootic lineages circulating since 2020, we found that a cow B3.13 isolate displayed enhanced replication in cow mammary gland cells, along with increased viral polymerase activity and stronger interferon antagonism in cow cells compared to an earlier EA-2020-C genotype virus. However, multiple avian and mammalian IAV strains, including other clade 2.3.4.4b high pathogenicity genotypes, were replication competent in bovine cells, particularly those of the mammary gland, suggesting that there is a diverse circulating IAV pool with the potential to infect cows. Moreover, we show that cow mammary cells co-express α-2,3 and α-2,6 – linked sialic acids, and are susceptible to co-infection with human and avian IAVs. We conclude that the US cow influenza outbreak does not simply reflect a unique adaptation of the B3.13 genotype virus; rather, the bovine udder represents a permissive niche for IAV and a plausible site for reassortment, underscoring its potential role in generating novel influenza viruses with pandemic risk.

Introduction

Since 2020, the H5N1 clade 2.3.4.4b of influenza A virus (IAV) has spread globally, threatening diverse wild bird species and causing economic devastation to the poultry industry. This panzootic has also resulted in multiple intra-phylum spill-over events, including farmed mink, semi-aquatic mammals and, in 2024, dairy cows [1, 2]). Since the detection of the B3.13 genotype H5N1 in cattle in March 2024, it has demonstrated sustained, widespread inter-state transmission in dairy cattle [3, 4]. This was unexpected, as until this event, fully verified evidence of IAV infection in cows has been limited, especially with regards to sequenced virus isolates available for further study [5]. While IAV is generally respiratory-tropic in mammals, the primary clinical indicator in the current bovine outbreak is an abrupt drop in milk production[6], correlating with high viral loads in milk[2]. Initial evolutionary and molecular analyses of cattle B3.13 viruses suggested that these viruses had quickly acquired mammalian associated adaptations in several genes [7]. Isolates were replication competent in human cell lines, were highly pathogenic in mice and ferrets [8, 9] and concomitantly, there have been multiple independent reports of zoonotic infections in dairy farm workers [10–12]. These events raised important questions around the likelihood of further IAV incursions into cattle and the pandemic potential of such outbreaks. We show here that a range of cow-derived cells are susceptible to infection with various IAV strains of avian, human or swine origin, that primary cow mammary cells express both α-2,3 and α-2,6 N-acetylneuraminic acid and that these cells can be co-infected with human and avian IAVs. We conclude that other IAV strains have the potential to infect cows and that this could yield reassortant viruses with zoonotic and pandemic potential.

Results

Cow mammary epithelial cells are susceptible to infection with diverse IAV strains

While IAV is traditionally considered to be a respiratory virus in mammalian species, transmission of B3.13 viruses between dairy cows has been associated with the use of milking equipment [13]. We therefore asked whether cow mammary gland epithelial cells are more susceptible to IAV infection than other cell types. Cow primary and immortalised cells originating from different anatomical sites (Figure 1A) were infected with a variety of IAV strains representative of four categories: laboratory-adapted, human, swine and avian. Infections were performed at low multiplicity of infection (MOI) and viral growth kinetic analyses were conducted up to 48 hours post infection (hpi). Of all tested cells, mammary cells showed the highest susceptibility to infection, with lab-adapted strains reaching viral titres of 108 PFU/ml. Titres of 106-7 PFU/ml were detected for viruses of human, swine and avian origins. Respiratory, anatomically diverse fibroblast, and immune cells were less permissive to IAV infection, with peak titres typically 2 to 3 orders of magnitude lower than those achieved in mammary cells across all virus strains (Figure 1B, Supplementary Figures S1-3). This further correlates with the ability of cow mammary cells to support the polymerase activity of diverse IAV strains (Figure 1C). The tropism of B3.13 for mammary gland epithelial cells therefore appears to be a generalisable phenomenon, observed across diverse IAV strains isolated from hosts spanning multiple taxonomic classes.

Cow cells are susceptible to infection with diverse IAV strains.

A). Schematic representation of the anatomical source of immortalised (green) and primary (black) cells used in this study. B) The indicated cells were infected with a variety of viral strains at MOI 0.1 (bMEC) or MOI 0.01 (for the remaining cells) and virus titres measured by plaque assay at 48 hpi. Viruses were subcategorised into lab-adapted, human, swine and avian strains as indicated. Circles represent individual data points from 3 independently performed experiments, except PBMC (n = 1). Lines represent median. Statistical annotations are the result of a 2-two-way ANOVA analysis. Tukey’s multiple comparisons tests were performed against MAC-T (amber asterisks) or bMEC (burgundy asterisks) data sets. C) RNP-reconstitution assays (minireplicons) were performed in MAC-T cells by transfecting pDUAL plasmids encoding segments 1, 2, 3 and 5 from the indicated viruses alongside a firefly luciferase-expressing vRNA-like reporter plasmid. Luminescence levels were measured at 48h post transfection. Data represent mean ± SEM from three independent experiments performed in triplicate.

B3.13 internal gene constellation confers better virus replication in cow cells

The inherent permissiveness of cow mammary cells to IAV infection does not explain why B3.13 was particularly successful at sustaining onward transmission in cows. We hypothesised that the unexpected dissemination of B3.13 infections in dairy cows could be explained by a fitness advantage of these viruses over other genotypes. To test this, biosafety level 2 viruses representative of an early B3.13 cow isolate (‘Cattle Texas’) and an avian precursor (AIV07, EURL genotype EA-2020-C) were generated harbouring the glycoproteins of the laboratory-adapted strain A/Puerto Rico/8/1934 H1N1 (PR8). Amino acid differences between the major internal gene products encoded by these viruses are summarised (Table S1). These viruses were used to infect bovine cells from different tissues including mammary epithelial cells (bMEC), type II pneumocytes, skin and heart fibroblasts and blood-derived macrophages (BDϕ) as well as mammary gland explants and surface-grown intestinal organoid cultures (2D enteroids). With the exception of bMECs, where AIV07 replicated slightly better, Cattle Texas replicated to 10 to 10,000-fold higher titres compared to AIV07 (Figure 2). Therefore, Cattle Texas demonstrated a fitness advantage in bovine cells compared to a representative avian virus precursor.

Cattle Texas is fitter in cow cells than an avian EURL genotype EA-2020-C precursor.

The indicated cow cell systems were infected with 2:6 reassortant viruses with PR8 glycoproteins and internal gene constellations of either AIV07 (blue) or Cattle Texas (yellow) (MOI 0.1 in B, 5000 PFU/explant in C and MOI 0.01 for all other systems). Supernatants were collected at the indicated times post infection and virus titres determined by plaque assay. Data represent mean ± SEM from two-four independent experiments each performed in duplicate. Statistical annotations represent the results of multiple unpaired t-tests performed for each individual time post infection. Cell types were immortalised mammary epithelial (MAC-T; A), primary mammary epithelial (bMEC; B), mammary gland explant (C), immortalised pneumocytes (BAT-II; D), skin fibroblast (E), cardiac fibroblast (F), blood derived macrophage (BDΦ) (G), and 2D enteroids (H).

After entering North America through the Atlantic bird migratory route, EA-2020-C viruses like AIV07 reassorted with the local low pathogenic avian influenza (LPAI) pool, giving rise to the B3.13 genotype via acquisition of new viral genome segments 1, 2, and 5 (encoding the PB2, PB1 and NP components of the viral ribonucleoprotein (RNP) complex respectively), [14] as well as segment 8 (encoding NS1 and NS2) [7, 15, 16] (Figure 3A). Therefore, we investigated whether the new RNP segments improved viral transcription activity in cow cells using a bovine-specific minigenome reporter assay [17]. Cattle Texas RNPs were 20-times more efficient than AIV07 (Figure 3B, open blue and yellow bars). To determine which segments contributed to this improved activity, Cattle Texas and AIV07 RNPs were reconstituted with reciprocal single gene replacements, including segment 3 (encoding PA), the final component of the viral polymerase. Replacing Cattle Texas PB1 with the AIV07 counterpart had no effect, while swapping NP increased RNP activity in MAC-T cells (Figure 3B, yellow cross-hatched bars). In contrast, PB2 or PA from AIV07 significantly decreased Cattle Texas minireplicon activity. In the reciprocal swaps, AIV07 polymerase activity was improved by up to 5-fold by Cattle Texas segments 1, 2 or 3, but replacement of segment 5 had an 8-fold negative impact on polymerase efficiency (Figure 3B, green cross-hatched bars). Thus, the increased polymerase activity of Cattle Texas is attributable to the PB2- and PA-encoding segments. To test this in a viral context, we made reassortant viruses in which Cattle Texas PB2 or all polymerase segments were replaced with those of AIV07. Both PB2 and full polymerase swap viruses exhibited reduced replication in cow mammary cells (Figure 3C), confirming the importance of these segments.

Increased fitness of Cattle Texas in cow cells correlates with enhanced viral polymerase activity and better NS1-driven inhibition of type I IFN induction.

(A) Schematic representation of AIV07 and Cattle Texas segment origins. Blue and yellow represent European and North American ancestry, respectively. (B) RNP-reconstitution assays (minireplicons) were performed in MAC-T cells by transfecting pDUAL plasmids encoding segments 1, 2, 3 and 5 from AIV07 and Cattle Texas in the indicated combinations alongside a firefly luciferase-expressing vRNA-like reporter plasmid. Luminescence levels were measured at 48h post transfection. Data represent mean ± SEM from three independent experiments each performed in triplicate. Statistical annotations are the result of one-way ANOVA analysis. Multiple comparisons were made against full AIV07 3PNP (blue asterisks) or Cattle Texas 3PNP (yellow asterisks) using a Dunnett’s test (C) MAC-T cells were infected with polymerase-segment reassortment viruses at MOI 0.01. Supernatants were collected at the indicated time points and virus titres measured by plaque assay. Data represent mean ± SEM from three independent experiments each performed in technical duplicate. (D) MAC-T cells were transfected with an IFN-β firefly luciferase-coding reporter alongside AIV07 or Cattle Texas segment 8 plasmids, and poly I:C treated 24h later. Luminescence readings were acquired after a further 24h. Data (left hand panel) represent mean ± SEM from three independent experiments each performed in technical triplicate. Statistical annotations are the result of one-way ANOVA analysis followed by a multiple comparison Dunnett’s test. A western blot (right hand panel) was also performed to detect NS1, using α-Tubulin as a loading control. (E) MAC-T cells were infected with segment 8 reassortant viruses at MOI 0.01. Supernatants were collected at the indicated time points and virus titres measured by plaque assay. For all shown statistical annotations: NS – non significant, *p-value <0.05, **p-value <0.01, ***p-value <0.001 and ****p-value <0.0001.

NS1 is the primary IAV IFN antagonist, working through multiple partially redundant but also strain-specific mechanisms, including both general repression of cellular gene expression at transcriptional and translational levels as well as specific inhibition of IFN induction [18]. As an initial test of whether reassortment of segment 8 was phenotypically important, we assessed host cell shut off by infecting MAC-T cells with the PR8 6:2 Cattle Texas or a PR8 2:5:1 Cattle Texas harbouring AIV07 segment 8 and pulsed the cells with puromycin at various times post infection. Western blotting for the puromycin-labelled prematurely terminated polypeptides indicated that while both viruses repressed cellular translation at late times post infection, there was no major difference between them (Figure S4A). As a more quantitative measure of shut off, we co-transfected MAC-T cells with segment 8 plasmids from the two viruses alongside a constitutive expression plasmid encoding Renilla luciferase and measured the resulting luminescence. Both NS1 proteins showed dose-dependent inhibition in this assay with Cattle Texas showing a modest improvement over AIV07 (Figure S4B). The PA-X accessory gene expressed from segment 3 also contributes to IAV host cell shut off [14]. Although segment 3 reassortment did not occur between genotypes EA-2020-C and B3.13, genetic drift led to four amino acid differences between the PA-Xs of AIV07 and Cattle Texas, three of which lie within the “X-ORF” (Figure S4C). Therefore, we similarly compared the ability of AIV07 and Cattle Texas PA-X proteins to inhibit host gene expression, but no significant differences were observed (Figure S4D).

To test specific IFN antagonism by NS1, the AIV07 and Cattle Texas segment 8 plasmids were transfected into MAC-T cells alongside an IFN-β reporter plasmid and the cells stimulated with poly I:C as a mimic for viral RNA. This showed that, at similar expression levels, Cattle Texas NS1 had a significantly higher ability to counteract type I IFN induction than AIV07 NS1 (Figure 3D). To assess the contribution of this to overall virus fitness, replication of the 2:5:1 Cattle Texas reassortant virus harbouring AIV07 segment 8 was characterised in MAC-T and BAT-II cells. Introduction of AIV07 NS into Cattle Texas resulted in a slight but non-statistically significant delay in virus propagation in mammary epithelial cells, but had a much larger effect in the pneumocyte cell line (Figure 3E), suggesting that changes in segment 8 also contribute to the improved fitness of Cattle Texas over AIV07. Thus overall, the reassortment events that replaced segments 1, 2 and 8 alongside drift adaptations in segment 3 may have contributed to the ability of the B3.13 genotype virus to infect cattle.

Spill-back of the cattle B3.13 virus into poultry has been observed on multiple occasions [3, 19], despite its evident fitness adaptations for bovines. The 2:6 Cattle Texas virus replicated to slightly lower titres in chicken lung epithelial cells than AIV07, while its polymerase and NS1 exhibited similar activities to those of the AIV07 proteins in the same cells (Supplementary Figure S5). Thus, the enhanced fitness of Cattle Texas in bovine cells is specific to bovine cells but does not carry a high fitness penalty in chicken cells.

Cow mammary cells are susceptible to European 2.3.4.4b genotype viruses and support co-infection by human and avian viruses

Detections of IAV in US dairy cattle are mostly of North American H5N1 2.3.4.4b genotypes B3.13 and, more recently, D1.1 viruses. Therefore, we asked whether similar dairy cattle incursions could happen in countries where the B3.13 and D1.1 genotypes are not circulating. To that end, we compared the growth kinetics of Cattle Texas with other avian European 2.3.4.4b genotypes in primary and continuous bovine mammary cells and explants, again all as 2:6 reassortants with PR8 glycoproteins to permit work at BSL2. In addition to AIV07 (genotype EA-2020-C), these included a 2.3.4.4b 2020 H5N8 precursor (H5N8-20) and EURL genotypes EA-2021-AB (AIV09) and EA-2022-BB (AIV48) (Figure 4A). In all three systems, Cattle Texas replicated to the highest titres, perhaps as expected for a virus with bovine-adapted internal segments. Other unadapted 2.3.4.4b strains showed varied viral fitness (Figure 4A). Cow mammary explants showed the greatest range of virus titres, with H5N8-20 and AIV07 exhibiting very poor fitness in this model, such that many of the explants were refractory to infection or only yielded low titres (Figure S6). However, the more recent AIV09 and AIV48 isolates replicated better in the explants, and Cattle Texas demonstrated the highest virus fitness, consistent with a trend for increasing replication fitness in mammary tissue of the 2.3.4.4b strain over time. This suggests that recent Clade 2.3.4.4b H5N1 genotypes on both sides of the North Atlantic have the potential to cause dairy cattle infections.

Cow mammary cells are infectable with viruses of human and avian origin.

(A) MAC-T (MOI 0.01), bMEC (MOI 0.1) cells and mammary explants (5000 PFU/explant) were infected with the indicated PR8 2:6 viruses. Supernatants were collected at the indicated times post infection and virus titres determined by plaque assay. Statistical analysis was performed by a 2-way ANOVA and multiple comparisons were performed against Cattle Texas in specific time points using a Dunnett’s test. (B) Annotated MALDI-TOF spectra of permethylated N-glycans from MAC-T and bMEC cells. Annotations show [M + Na]+ molecular ions. Peak annotation is based on composition, biosynthetic knowledge and MS/MS analysis. (C) Proportions of a-2,3 and a-2,6 linked sialic acid in MAC-T and bMEC N-glycans. (D) MAC-T cells were infected with combinations of Norway 2018 (pH1N1) and/or Belgium H3N1 at MOI 5 for 16h. Cells were fixed, stained with anti-H1 and H3 before quantification was performed by FACS analysis. Data represent mean ± SEM from three independent experiments each performed in technical duplicate. For all shown statistical annotations: ****p-value <0.0001.

Human IAV pandemics classically arise from reassortment between avian and mammalian strains [20, 21]. Historical data suggest the incursion of human seasonal strains into cattle [5], so in the face of an ongoing H5N1 epidemic in birds and cattle, the possibility of the cow udder as a “mixing vessel” arises. However, co-infection of a single cell harbouring both avian (α-2,3 linked sialic acid) and mammalian (α-2,6 linked sialic acid) receptor types is needed for this mixing to occur. Lectin staining analyses indicate the presence of α-2,3 linked sialic acid receptors for circulating avian H5 viruses in cow mammary glands [22–24], while our own infectivity data (Figures 1, S1) suggest the presence of human α-2,6 linked sialic acids on bovine mammary epithelial cells. To establish the amounts of each receptor type in the cells used here, we next performed quantitative glycomic analyses on the immortalised and therefore likely relatively homogeneous (MAC-T) and primary, potentially heterogeneous (bMEC) cell populations. This identified complex N-glycans containing a mix of α-2,3 and α-2,6 linked sialic acids, with a higher proportion of α-2,6 linkages present on the bMECs compared to MAC-T cells (Figure 4B, C; Supplementary Figures S7-S10). Cow mammary cells therefore express the sialic acids required for infection with both avian and mammalian IAVs; however, this does not demonstrate that the two sialic acid conformations are present at functionally relevant levels on the same cell. To determine whether individual cow mammary gland cells could be co-infected by human and avian strains of virus, bMEC cells were inoculated with an α-2,6 sialic acid-dependent human seasonal H1N1 strain and an α-2,3 sialic acid-dependent low pathogenicity avian virus, either singly or together. Infected cells were fixed and stained with anti-H1 or H3 HA and analysed by flow cytometry. The individual infections confirmed antibody specificity, while a dual-infected population was clearly visible in the co-infection sample (Figure 4D). We therefore conclude that co-infections of human and avian viruses within the cow mammary glands are possible, supporting the possibility of reassortment events.

Discussion

The unexpected influenza outbreak in US dairy cows requires careful consideration to understand why it occurred, and its potential implications for both livestock and human health [15]. At one extreme, it could have resulted from a rare adaptation of the B3.13 genotype to allow infection of a species usually refractory to infection with most strains of IAV; at the other, an unusually large and well documented epidemic of a not uncommon spillover event that mostly goes undetected. To experimentally test these possibilities, we compiled a variety of bovine cell culture systems and performed a systematic analysis of their susceptibility to IAV infection. We found that while there are clear virus strain and cell-type differences in the susceptibility of bovine cells to IAV infection, many cell types are susceptible to infection with a diverse range of avian and mammal viruses. Notably however, we think our data support mammary gland epithelial cells to be the primary site of infection compared to respiratory, immune or fibroblast cells, consistent with the clinical manifestations of mastitis and high viral loads present in milk [4, 25]. The bovine cells used in our study come from a mixture of dairy and beef cattle. The mammary explants were derived from dairy cattle of unspecified breeds, while MAC-T cells are from Holstein Friesian dairy cattle. The bMECs were derived from beef cattle, and the enteroids from male Holstein Friesian calves. We can therefore be confident that in vitro cell permissivity to IAV infection is not unique to a single breed, sex or age of cattle, but we cannot rule out that these factors might influence viral susceptibility in vivo [26].

The ability of cow mammary cells to be infected with and support high levels of replication of a wide range of IAV strains here described argues against the extreme hypothesis of a truly unique phenotypic trait of the B3.13 strain that permits infection of cows. These observations are consistent with recent studies showing that other avian 2.3.4.4b isolates are able to replicate in bovine cells [27–29] and udders [25, 30], as well as the more recently detected D1.1 genotype spillover into US dairy cattle [31]. However, our findings of wide variation in the ability of specific strains of IAV to replicate in bovine cells suggests that adaptation does play a role. Thus, we sought to determine what, if anything, predicted the emergence and transmission of Cattle Texas above other avian IAV strains. We found a multifactorial scenario, with improvements in polymerase efficiency and the ability of NS1 to counteract type I interferon induction. These improvements were apparent in cow but not chicken cells, indicative of evolutionary adaptations to the new host. This is consistent with other studies showing a replicative advantage of cattle isolates over an avian precursor [8, 30]. For PB2 and PA, the amino acid mutations PB2 M631L and PA K497R present in the Cattle Texas strain we used have recently been reported to enhance polymerase efficiency and virus replication in cow mammary cells [3, 17, 19]. Furthermore, impairment to transcriptional efficiencies rendered by the newly acquired segment 5 imply that there is still the potential for further fitness gains of the B3.13 genotype in cattle. Notably, the D1.1 strain isolated from cows in Nevada possess PB2 D701N [32] while a human isolate contained PB2 E627K [33], suggesting that other adaptation pathways in cattle are possible [34, 35].

Pigs are the classic reassortment host of IAV, the respiratory tract of which is unusual in its ability to support infection with both avian and mammalian IAV strains and from which the most recent influenza pandemic in 2009 arose [36, 37]. However, HPAI H5 infections in pigs are rare; pigs were either not susceptible to H5N1 2.3.4.4b strains following nasal exposure [38], or when infected no transmission was observed [39]. Therefore, reassortment of H5 HPAI with other mammalian IAV strains within the swine respiratory tract can be considered unlikely. However, our observations on the ability of avian and mammalian IAV strains to infect cow mammary glands raised the question of whether cow udders could serve as tissues permissive for viral reassortment for the generation of novel strains containing human-adapted and H5 HPAI-origin segments, analogous to the pig reassortment events that are conjectured to have precipitated previous pandemics. The fact that primary cow mammary cells can be co-infected with human and avian IAV strains, supported by the mixed distribution of α-2,3 and α-2,6 linked sialic acid observed in these cells, suggests that cow udders do have this potential. However, our experiments necessitated culture of these cells and culture passage has been observed to increase overall sialylation in some cell types [40, 41], meaning that our results may modestly overestimate the total sialylation abundance.

Infection of cows with human and other IAV strains has been suggested in the past. In 1997, dairy cows in the southwest of England presented with sporadic drops in milk production, which the reporting authors considered to be attributable to IAV [42, 43]. Additional studies have reported seroconversion against circulating seasonal H3N2 virus in British cattle, which was associated with reduced milk yield and respiratory disease [44, 45]. Despite the occasional respiratory signs, such as nasal discharge and increased respiratory rate followed by seroconversion, virus isolation from respiratory swabs from these seroconverted animals was unsuccessful. With hindsight, isolation from milk may have had a higher chance of success. Consequently, the idea that “cows could foster flu pandemics” was considered back in the early 2000s [46]. Moreover, the recent seroconversion for H5N1 2.3.4.4b in sheep and detections of IAV RNA in milk [47, 48] further suggests that other ruminant livestock species may hold similar potential. Until recently, the absence of H5 detection in cattle beyond the United States might have indicated that the coordinated polymerase and NS1 adaptations that we have linked to this type of spillover have not yet emerged in Europe or other regions. It could also be attributable to insufficient surveillance, potentially due to the unexpected presentation as mastitis. However, H5 seropositive dairy cattle in the Netherlands have recently been reported [49]. At the time of writing, no viral sequences have been made available, so whether this represents similar viral adaptation to what we describe for the B3.13 genotype and/or enhanced surveillance remains to be determined.

Overall, our findings point to requirements for reconfigured and enhanced policy procedures and surveillance of IAV incursions into ruminant livestock species. Currently, mastitis outbreaks are not routinely investigated for the presence of IAV. Adding IAV testing to mastitis diagnostic assessments as a differential when other common causes have been excluded will be beneficial to livestock. This will allow early detection of emerging events with implementation of appropriate infection control measures and justify contact tracing of farm workers.

Materials and Methods

Ethics

All reverse genetics work used a loss-of-function approach and was carried out under a license (GMRA1811) from the UK Health & Safety Executive. All work with avian viruses was carried out in a segregated BSL2 laboratory in the Roslin Institute, separate from where mammalian IAV strains are handled.

All animal experimental work was completed under license and in accordance with the Animal (Scientific Procedures) Act 1986 and with approval from the Roslin Institute animal welfare and ethics committee under the project licence number PP0612543. All studies were performed in line with the Animal Research: Reporting of In Vivo Experiments (ARRIVE) Guidelines [50].

Cell line maintenance, primary cell isolation, derivation and culture

Madin-Darby Canine Kidney (MDCK) (American Type Culture Collection, ATCC), MDCK expressing galline ANP32A (MDCK-GgANP32A) [51], MDCK expressing Sialyltransferase 1 (MDCK-SIAT) [52], human embryonic kidney (HEK293T) and immortalised mammary epithelial cells (MAC-T) [53] were cultured in DMEM supplemented with 10% foetal bovine serum (FBS), 2mM glutamine, 100 U/ml penicillin, and 100 μg/ml streptomycin. MDCK-SIAT cells were further supplemented with 50 µg/ml geneticin. MAC-T cells were further supplemented with 5 µg/ml insulin and 1 µg/ml hydrocortisone. Chicken lung cells (CLEC213) [54] were cultured with DMEM/Nutrient Mixture F-12 also supplemented with 8% FBS, 100 U/ml penicillin and 100μg/ml streptomycin. BAT-II cells [55, 56] were maintained in Small Airway Growth medium (Lonza) further supplemented with 10% FBS, 100 U/ml penicillin and 100μg/ml streptomycin. All immortalised cells were regularly tested for mycoplasma contamination and passaged once or twice weekly.

Primary cow mammary epithelial cell (bMEC) lines generated from the mammary glands of Aberdeen Angus cross cows as previously described [57]. Glands were sprayed with 70% ethanol and skin was removed. Cuboidal epithelial tissue fragments (5 cm in length) were cut from the gland, submerged in 70% ethanol for 30s and washed 5 times in PBS containing antibiotics and 2.5 μg/ml amphotericin B. Tissue fragments were transferred to a glass petri dish and minced with scalpels to 2-3 mm pieces. Next, 10-15g of minced tissue was dissociated overnight with low shaking in 20 ml of DMEM/Nutrient Mixture F-12 supplemented with 15 mM HEPES and 0.075% sodium bicarbonate, 2% Bovine serum albumin (BSA), 300 U/ml Collagenase Type I and 100 U/ml Hyaluronidase from bovine testes. Dissociation media was removed by centrifugation and pellet was washed twice with DMEM/F12. This yielded small aggregates of part-digested tissue which were resuspended in freezing medium and stored at −150°C. To obtain a single cell suspension, aliquots of part-digested tissue aggregated were thawed, washed with Hanks’ Balanced Salt Solution (HBSS) supplemented with 2% FBS, resuspended in warm trypsin and vigorously pipetted for 3 mins and washed with HBSS/2% FBS, resuspended in a warm Dispase/DNAse solution and gently pipetted for 3 minutes. Ice cold HBSS/ 2% FBS was added and the suspension was filtered on a 70 μM strainer, cells were washed by centrifugation, resuspended in growth media (DMEM/F12 supplemented with 0.1% BSA, 10 ng/ml epidermal growth factor (EGF), 10 ng/ml cholera toxin, 1 µg/ml bovine insulin, 0.5 µg/ml hydrocortisone, 100 U/ml penicillin, 100 μg/ml streptomycin and 250 µg/ml gentamycin) and seeded onto collagen coated wells/flasks. bMECs were examined everyday by light microscopy to check for signs of contamination, viability, morphology, and confluence. Media was refreshed daily. Growth media was further supplemented with 2.5 µg/ml amphotericin B, the concentration of which was halved every day until the cells had reached sufficient confluency to be expanded.

Primary fibroblasts were extracted from skin and brain tissue explants dissociated with 5 mg/ml Collagenase type I at 37°C, shaking for 3h. The dissociated material was passed through a 70 µm filter and washed with HBSS. Cells were cultured on bovine gelatine-coated tissue culture plastic and expanded in Glasgow’s MEM) supplemented with 10% foetal bovine serum, 2 mM L-glutamine, 1x MEM non-essential amino acids, 1 mM sodium pyruvate, 0.1 mM 2-mercaptoethanol, 100 U/ml penicillin-streptomycin, 50 µg/ml gentamicin and 2.5 µg/ml amphotericin B. Established cultures were maintained in the same medium without antibiotics.

Cow PBMCs were collected from Holstein-Friesian cattle housed at the University of Edinburgh farm using published procedures [58]. Cow blood derived macrophages (BDΦ) were generated from PBMCs as described previously [58] except that the macrophages were cultured in BDΦ medium (RPMI-1640 supplemented with 20% FBS, 4 mM L-glutamine and 50 µM β-mercaptoethanol, 100 U/ml penicillin-streptomycin and 20 ng/ml His-tagged recombinant bovine CSF1 - the latter supplied by the Roslin Immunological Toolbox). BDΦ medium was replaced on day 7. On day 12 the adherent cells were rigorously washed with phosphate-buffered saline PBS and detached with TrypLE Express. Purified BDΦs were resuspended at 1.5 x 105 cells/ml in BDΦ medium without CSF1, aliquoted into 24 well plates (1ml/well) and cultured for 48h before infection on day 14.

Bovine 3D enteroids were derived from five 6-month-old healthy male British Holstein–Friesian calves housed at the University of Edinburgh farm. Calves were subject to Schedule 1 method of cull and sections of distal ileum were obtained post-mortem and stored in HBSS on ice until further use. Crypts were isolated and 3D enteroids were cultured as previously described [59]. Bovine 3D organoid cultures were used to generate 2D enteroids as previously described [60].

Bovine mammary gland explants were extracted from udder tissue collected at Sandyford Abattoir (Paisley, Scotland) from healthy lactating cows of unspecified breeds that had been culled for the food chain. Excess tissue was removed, teat and gland cistern were dissected and transferred to transport media comprising of chilled DMEM supplemented with 100 μg/ml Penicillin/ 100 μg/ml streptomycin and 5 μg/ml fungizone. Every 45 minutes, tissues were transferred to 500 ml of fresh transport media. Inside class II biological safety cabinet, explants were made using a 5 mm biopsy punch. Biopsies were added to 24 well plates with 500ul of transport media further supplemented with 10% FBS and incubated for 24h at 37°C, 5% CO2 prior to infection.

Validation of bovine origin of the primary and continuous cell lines was performed by PCR. DNA was extracted from cell pellets, and the cytochrome B gene was amplified by PCR with degenerate primers (CGAGACGTVAAYTAYGGMTGA and ACBGTDGCYCCTCARAADGATATTTG). Cycling conditions were 94°C for 2 minutes, followed by 50 cycles of 94°C for 18 seconds, 50°C for 21 seconds then 72°C for 30 seconds with a final extension of 72°C for 5 minutes. Amplicons were sent for Sanger sequencing and resultant sequences were aligned to the Bos taurus mitochondrion complete genome reference sequence (NCBI accession number NC_006853.1).

All cells, organoids and explants were kept at 37°C and 5% CO2 under normoxic conditions.

Viruses

Selected viruses were split into 4 different categories based on their history and isolation animal host (Table 1). A/Puerto Rico/8/1934 H1N1 (“PR8”) [61], mouse adapted A/California/04/2009 H1N1 (“Cal04”) [62], A/WSN/33 H1N1 (“WSN”), swine adapted A/mallard/Netherlands/10-Cam/1999 H1N1 (“Mallard”) [63, 64], A/Udorn/307/1972 H3N2 (“Udorn”), A/England/195/2009 pH1N1 (“Eng195”), A/swine/England/87842/1990 H3N2 (“Swine H3N2”) and 2:6 reassortants of 2.3.4.4b H5N1 viruses containing PR8 segments 4 and 6 viruses were generated by reverse genetics as previously described[65]. Briefly, 2 × 106 HEK293T cells were transfected in OptiMEM with pDUAL reverse genetics plasmids (250 ng of plasmid for each virus segment), using 4 µl of Lipofectamine 2000 according to the manufacturer’s instructions. At 24 hours post transfection, the media was replaced by serum-free DMEM supplemented with 0.14% (w/v) fraction V BSA and 1 μg/ml of l-(tosylamido-2-phenyl) ethyl chloromethyl ketone (TPCK)-treated trypsin. Virus-containing supernatant was collected after a further 2-days incubation (here described as a “P0 stock”). P0 stocks of PR8, Swine H3N2 and Udorn were propagated in MDCK cells. P0 stocks of Mallard, Eng195 and PR8 2:6 2.3.4.4b H5N1 reassortants were further propagated in 10-day embryonated hen’s eggs. Allantoic fluids were collected 48hpi. Cal04 P0 stocks were propagated in MDCK-SIAT cells. Reassortants between A/dairy cattle/Texas/24-008749-001-original/2024 H5N1 (“Cattle Texas”) and A/chicken/England/053052/2021 H5N1 (“AIV07”) were propagated in MDCK-GgANP32A.

Isolates of A/USSR/90/1977 H1N1 (“USSR”) A/Brisbane/10/2007 H3N2 (“Brisbane”), A/Norway/3433/ 2018 H1N1 (“Norway”) and A/swine/453/England/2006 H1N1 (“Swine H1N1”) isolates were propagated in MDCK cells. Clarified supernatants were typically collected at 36–72 hpi. Isolates of A/chicken/Belgium/460/2019 H3N1 (“H3N1”), A/chicken Belgium/150/1999 H5N2 (“H5N2”), A/chicken/Netherlands/917/2010 H6N1 (“H6N1”) and A/chicken/Saudi Arabia/2525/2000 H9N2 (“H9N2”) isolates were propagated in hen’s eggs and allantoic fluids were collected and clarified at 48 hpi. All viruses were sequenced to confirm identity using Nanopore sequencing (described below).

RNA extraction and Multi-segment RT-PCR

To confirm the genome sequences of the virus strains used, viral RNA was extracted from cell- and egg-grown viral stocks using the QIAamp Viral extraction mini kit (QIAGEN) as per manufacturer’s instructions. Viral genomes were amplified using the Superscript III One-Step RT-PCR kit (Invitrogen), 5 µl of extracted viral RNA, GGGGGGAGCAAAAGCAGG and GGGGGGAGCGAAAGCAGG as forward and GGGGTTATTAGTAGAAACAAGG as reverse primers[66]. Thermocycling conditions for RT-PCR reaction were: 55°C for 2 mins, 42°C for 60 mins, 94°C for 2 mins; 5 cycles of 94°C for 30 s, 44°C for 30 s and 68°C for 3 mins 30 s; 26 cycles of 94°C for 30 s, 57°C for 30 s and 68°C for 3 mins 30 s and 10 mins hold at 68°C. Resulting PCR was analysed on an 1% agarose gel.

Nanopore sequencing and data analysis

RT-PCR products were pooled and prepared for sequencing using a ligation sequencing amplicon kit (SQK-NBD114.24, Oxford Nanopore Technologies, Oxford, UK) and sequenced on a FLO-MIN114 flow cell (R10.4.1., Oxford Nanopore Technologies) on an in-house GridION device. Final library concentration was approximately 10 fmol. Molarity was determined using Oxford Nanopore recommended tables, assuming an average multi-segment RT-PCR fragment size of 1.3 kb. Reads were called using the high accuracy basecall setting on the MinKNOW UI (Oxford Nanopore Technologies). The “passed” FASTQ output files were used in the sequencing analysis. Data was analysed using the EPI2ME (Oxford Nanopore Technologies) and Sequeduct (Edinburgh Genome Foundry) [67] pipelines. The processed FASTQ files were aligned against expected sequences. For the Sequeduct pipeline, reference sequences were converted to Genbank format. Consensus sequence output from the “analysis” workflow was used to align against expected sequences.

Plaque assays

Confluent monolayers of MDCK cells (1 × 106 cells seeded in 12-well plates the day prior to infection) were washed once with PBS, then infected with 250 µl of tenfold serial dilutions of virus. Following a 1 h incubation at 37 °C that allowed virus adsorption, cells were overlaid with DMEM supplemented with 0.14% fraction V BSA, 1 µg/ml TPCK-treated trypsin and 1.2% Avicel. Following a 48h undisturbed incubation at 37 °C, cells were fixed with 10% Neutral Buffered Formalin for at least 20 minutes and stained with a 0.1% toluidine blue solution for at least 1 h and plaques were then counted. Plaque assays for avian viruses or PR8 2:6 viruses with avian origin PB2 were performed as described but using MDCK-GgANP32A cells. Plaque assays for human viruses Eng195, Cal04, Brisbane and Norway H1N1 were performed using MDCK-SIAT cells and incubation temperatures of 35 °C for 72h.

Virus replication kinetics

Monolayers of cells were washed once with PBS and infected with virus diluted in serum-free medium for 1 h at 37 °C (the multiplicity of infections used is specified in each figure legend). Infectious medium was replaced with serum-free medium supplemented with 0.14% fraction V BSA and 1 µg/ml TPCK-treated bovine pancreas trypsin.

Individual mammary explants were transferred to 96-well plates and infected with 5000 PFU of virus in 100 µl of serum free DMEM for 2h at 37°C. Explants were transferred to clean 24-well plates, washed 3x with warm PBS and overlayed with 1ml of DMEM supplemented with 0.14% fraction V BSA and 1µg/ml TPCK. Supernatant was collected at different timepoints post infection and infectious titres were determined by plaque assay.

Immunoblotting

Cells were lysed in Laemmli buffer, heated at 80 °C for 5 min and subjected to polyacrylamide gel electrophoresis using sodium dodecyl sulfate-based buffer. Peptides were transferred to polyvinylidene difluoride membranes for 90 minutes at 32 V. Membranes were blocked for 1 h with 1× Tris-buffered saline with 0.1% Tween20 (TBST)/5% milk, washed three times with TBST and stained with primary antibodies diluted in TBST/5% BSA overnight at 4 °C. After three 5-minute TBST washes, secondary antibodies were diluted in blocking buffer and incubated for 1 h at room temperature protected from direct light. Following three more washes in TBST, membranes were imaged using the LI-COR Fc-Odyssey Imaging platform using the Image Studio Lite Software.

Plasmid-based IFN- β promoter reporter assays

Under confluent monolayers of MAC-T or CLEC213 cells were co-transfected with 500ng or 100 ng of reporter plasmid encoding firefly luciferase under the control of human [68] or chicken IFN-β [69] promoter and 1000ng of segment 8 pDUAL plasmids using 2 μl of Lipofectamine 2000. 24 hours post transfection, cells were stimulated with 5 μg of Poly I:C by transfection using 2 μl of Lipofectamine/well. Mock-stimulated cells were also included where Poly I:C was replaced by water. 24 hours post-stimulation, cells were lysed with 100 μl of reporter lysis buffer and luciferase activity was measured from 40 μl of lysate in opaque 96-well plates by using 25 μl of luciferase assay reagent in Cytation3 Plate Reader and Imager (integration time 1 second).

Cellular gene expression shutoff transfection-based assays

Sub-confluent monolayers of MAC-T and CLEC213 cells were co-transfected in triplicate with a range of pDUAL segment 3 or 8 effector plasmids and 200ng of pRL using 2μL of Lipofectamine2000. Forty-eight hours post transfection, cells were lysed and Renilla luciferase expression was quantified using a Renilla Luciferase Assay kit. Briefly, cells were lysed with 100μL of 1× lysis buffer and frozen overnight at −20°C. Cells were scraped off the plate, cell debris was removed by centrifugation (8000rpm, 5 minutes, 4°C). Luminescence was measured from the mix of 20 μl of lysate and 50 μl of 1× Renilla luciferase substrate using a Cytation3 Plate Reader and Imager (integration time 1 second).

Puromycin-labelling of nascent proteins

Subconfluent monolayers of MAC-T cells were infected at MOI 5. At 9, 12, 15 and 18 hpi, supernatant was replaced with fresh medium supplemented with 20 μg/ml of puromycin dihydrochloride for 1h. Cells were washed twice with PBS, lysed in Laemmli buffer and subjected to immunoblotting using a primary anti-puromycin antibody (Millipore; MABE343).

RNP reconstitution (minireplicon) assays

Sub-confluent monolayers of MAC-T or CLEC213 cells (1 × 105 cells seeded in 24-well plates the previous day) were co-transfected in triplicate with 50 ng of each pDUAL plasmid encoding PB2, PB1, PA and NP along with 20 ng of a bovine PolI-driven vRNA-like firefly luciferase reporter plasmid [17]. As negative controls, transfections lacking the PB2 plasmid (with empty pDUAL vector used to balance plasmid amount) were also performed. Two days post transfection, cells were lysed with 120 µl of reporter lysis buffer (Promega E4030). Lysates were cleared and luminescence was measured from 40 μl of lysate in opaque 96-well plates by using 25 μl of luciferase assay reagent in a Cytation3 Plate Reader and Imager (integration time 1 second).

Tissue homogenisation

Mammary explants were added to 1ml of serum-free DMEM and a 5mm diameter stainless steel bead in 2ml tubes and homogenised using a Tissue Lyser II (Qiagen) system at 28kHz for 3 sets of 2 minutes. Homogenised tissue was clarified by centrifugation (3000 rpm, 5 min, 4°C). Homogenates were titrated for infectious virus by plaque assay.

Glycomic analysis of MAC-T and bMEC cells

Approximately 10 million cultured MAC-T and bMEC cells were washed and scraped into cold PBS. Cells were then pelleted and washed several times with cold PBS prior to processing[70]. Briefly, cells were lysed by sonication, dialysed and lyophilised. Reduction and carboxymethylation reactions were carried out on the lyophilised material, which was subsequently dialysed and lyophilised. Samples were digested with trypsin (Sigma) and the resultant glycopeptides purified by Sep-Pak C18 chromatography. N-glycans were released by PNGase-F (Roche) and purified by Sep-Pak C18 chromatography. N-glycan samples were then split into three aliquots. Two of the three aliquots were treated with either sialidase-S or sialidase-A (Agilent Technologies) and then purified by C18 Sep-Pak chromatography, the third aliquot was untreated. Untreated and treated glycans were then permethylated and analysed using a 4800 MALDI-TOF mass spectrometer (Applied Biosystems).

MS and MS/MS data were visualised using Data Explorer (Applied Biosystems) and [M + Na]+ molecular ions were annotated manually with the assistance of GlycoWorkbench [71]. Glycan structures were predicted based on composition and knowledge of biosynthetic pathways, with MS/MS analysis used to confirm assignments. Figures were prepared using Adobe Illustrator.

Glycomic data was quantified using an R programme described in Wu et al., 2024 [72] which detects glycans according to their isotopic peak pattern and calculates the relative intensity across the isotope cluster. An R programme was then used to calculate the proportions of a-2,3 and a-2,6 sialic acid. First, the relative intensity of each glycan was normalised to the sum of the relative intensities of all the glycans in the sample. For sialylated glycans, normalised intensities were then weighted based on the number of sialic acid residues present on each glycan. The proportion of a-2,6 linked sialic acid was determined by summing the weighted normalised intensities of sialylated molecular ion peaks remaining after sialidase-S digestion. The proportion of a-2,3 linked sialic acid was calculated by subtracting the weighted normalised intensities of sialylated molecular ions in the sialidase-S sample from the weighted normalised intensities of sialylated molecular ions from the untreated sample (raw data in Supplementary File) [73].

Cell staining and flow cytometry

Infected bMECs were trypsinised, resuspended and fixed in PBS/4% formaldehyde. Following a PBS wash by centrifugation, cells were blocked with PBS/2%BSA for 1h. Cells were stained with rabbit anti-H3 (c33) or mouse anti-H1 (2-12C, Absolute antibody) primary antibodies diluted 1:5000 in blocking buffer for 1h at room temperature and with constant agitation. Cells were then washed twice 2x with blocking buffer and resuspended in a secondary antibody solution containing donkey anti-rabbit647 (ab150075, abcam) and donkey anti-mouse 488 (A11011, Invitrogen) for another 30 min. Cells were centrifuged, washed twice with blocking buffer and analysed using a BD LSR Fortessa X20 cytometer.

eLife Assessment

This study presents important findings that bovine mammary epithelial cells can be infected with both avian and human influenza A viruses, providing a potential site for viral reassortment. The evidence to support these claims is generally solid; however, the evidence suggesting lower permissiveness of cells from other organs is incomplete. The work will be of interest to virologists and evolutionary biologists working on cross-species transmission of viruses and pandemic preparedness.

Reviewer #1 (Public review):

Summary:

Here, Pinto and colleagues set out to investigate whether the cow udder is a potential mixing site for the influenza virus. The authors have demonstrated that bovine mammary epithelial cells can be infected with both avian and human influenza A viruses, supporting the idea that the cow udder may be a potential site for reassortment. Furthermore, they demonstrate that the bovine-adapted IAV replicates to similar titers in avian epithelial cells when compared to an AIV precursor virus. Thus, suggesting there is no fitness trade-off, and confirms the potential for spill-back of the cattle B3.13 into poultry, which has already been observed. Overall, I believe the authors achieved their aims. However, there are instances in which the results do not entirely support the conclusions (noted in weaknesses). Given the ongoing questions surrounding highly pathogenic avian influenza A virus in dairy cows, this work provides valuable evidence for the potential of the cow udder as a site of reassortment. These findings highlight the need for surveillance of influenza A virus incursions into livestock species, particularly cows. Some specific strengths and questions regarding weaknesses have been outlined below.

Strengths:

(1) The authors use a diverse range of cell types and influenza A virus strains, as well as a wide range of techniques to address the questions at hand.

(2) The use of cells from multiple bovine breeds for the MAC-T, bMEC and explants suggests the phenomenon is not unique to a single breed.

(3) The results suggesting there is no fitness trade-off for Cattle Texas in an avian host are interesting, and confirm the potential for spill-back of the cattle B3.13 into poultry, which has been observed.

Weaknesses:

I have listed my complete questions/concerns below. However, there are two main weaknesses of the article in its current state. Firstly, there is no apples-to-apples comparison in terms of determining a preference for IAV to infect the cow udder over other organs (Q4). The mammary gland and respiratory tract are represented by epithelial cells, but for other organs, fibroblasts were chosen. I think the fairer comparison would be to compare epithelial cells from different organs to demonstrate a preference for the mammary gland. Secondly, the main premise of the article relies on bMEC and MAC-T (primary and immortalised mammary epithelial cells), facilitating higher viral growth than the cells from other organs. Yet throughout the article, a 10x higher dose of IAV is used in the bMEC cells compared to everything else (Q6). This raises the question of how much of the results are due to a preference for the mammary epithelial cells, and how much is simply due to the increased dose.

Reviewer #2 (Public review):

The authors use a library of influenza A viruses from different strains, classified in lab-adapted, human, avian, and swine according to the animal from which they were isolated. They propose that the cow mammary gland serves as a mixing vessel for influenza A viruses. As a first approach, the authors assess susceptibility to infection across different cell types, including continuous and primary cell lines, bovine mammary cells, and mammary explants. All these cells support polymerase activity. Then, they analyzed changes in the bovine virus's viral fitness relative to an avian precursor. The authors use single-gene replacement to study whether and which RNP segments improve viral transcription. As part of this section, they also test IFN-specific antagonism by NS1 to assess the input of segment 8. Quantitative glycomic analysis was performed on the continuous bovine mammary cell line to demonstrate the presence of both a2,3 and a2,6, which is consistent with their observation that these cells can be co-infected with human and avian IAVs simultaneously. The main question, however, is: what is the glycome in the explants, or directly from tissues?

Overall, the manuscript is clearly written and provides new insights into the behaviour of the cattle isolate, now compared with a representative group of model or precursor HAs of different origins.

It would be great if a consistent nomenclature for the IAV strains could be used in the study. There is a mix of origin (Texas), animal from which the virus was isolated (mallard), or abbreviations that do not follow guidelines (IAV07). Are the USSR and Udorn not lab-adapted?

The experimental setup includes bovine mammary primary and continuous cells, as well as mammary explants. Some of the most significant differences, for example, in viral fitness studies and co-infection experiments, are observed in these explants. Perhaps there could be some additional focus on this observation. The implications in comparison to the results obtained in cultured cells could be described. How will the human and other HA subtype viruses fare in the explants?

Reviewer #3 (Public review):

Summary:

This excellent manuscript by Pinto, Sharp, and colleagues examines bovine tissue tropism for influenza viruses. They find that bovine flu, as well as other strains, has strong replication in mammary tissue. They also map the genetic changes to influenza that improve replication in bovine cells. Overall, the study is well designed and executed, and the results are very timely.

Strengths:

(1) The experiments are well-controlled.

(2) The figures are well-constructed and easy to follow.

(3) The Methods and legends are detailed, with sufficient information.

Weaknesses:

(1) A comparison to human cells would strengthen the overall impact of the results. Are human mammary cells also uniquely susceptible to influenza? Are bovine mammary cells special in some way?

(2) For the virus infection studies with segment 8 swaps, it should at least be noted that some of the phenotypes could be driven by NEP.

(3) The data demonstrating that bMEC can support co-infection are compelling and important, but would be strengthened with a comparison from a different cell type or species. Do mammary cells uniquely support higher co-infection?

Author response:

The following is the authors’ response to the original reviews.

Public Reviews:

Reviewer #1 (Public review):

Summary:

Here, Pinto and colleagues set out to investigate whether the cow udder is a potential mixing site for the influenza virus. The authors have demonstrated that bovine mammary epithelial cells can be infected with both avian and human influenza A viruses, supporting the idea that the cow udder may be a potential site for reassortment. Furthermore, they demonstrate that the bovine-adapted IAV replicates to similar titers in avian epithelial cells when compared to an AIV precursor virus. Thus, suggesting there is no fitness trade-off, and confirms the potential for spill-back of the cattle B3.13 into poultry, which has already been observed. Overall, I believe the authors achieved their aims. However, there are instances in which the results do not entirely support the conclusions (noted in weaknesses). Given the ongoing questions surrounding highly pathogenic avian influenza A virus in dairy cows, this work provides valuable evidence for the potential of the cow udder as a site of reassortment. These findings highlight the need for surveillance of influenza A virus incursions into livestock species, particularly cows. Some specific strengths and questions regarding weaknesses have been outlined below.

Strengths:

(1) The authors use a diverse range of cell types and influenza A virus strains, as well as a wide range of techniques to address the questions at hand.

(2) The use of cells from multiple bovine breeds for the MAC-T, bMEC and explants suggests the phenomenon is not unique to a single breed.

(3) The results suggesting there is no fitness trade-off for Cattle Texas in an avian host are interesting, and confirm the potential for spill-back of the cattle B3.13 into poultry, which has been observed.

Weaknesses:

I have listed my complete questions/concerns below. However, there are two main weaknesses of the article in its current state. Firstly, there is no apples-to-apples comparison in terms of determining a preference for IAV to infect the cow udder over other organs (Q4). The mammary gland and respiratory tract are represented by epithelial cells, but for other organs, fibroblasts were chosen. I think the fairer comparison would be to compare epithelial cells from different organs to demonstrate a preference for the mammary gland. Secondly, the main premise of the article relies on bMEC and MAC-T (primary and immortalised mammary epithelial cells), facilitating higher viral growth than the cells from other organs. Yet throughout the article, a 10x higher dose of IAV is used in the bMEC cells compared to everything else (Q6). This raises the question of how much of the results are due to a preference for the mammary epithelial cells, and how much is simply due to the increased dose.

(Q4) When we set out to test if cow mammary gland cells were particularly susceptible to IAV infection compared to other bovine cell types, we used what was available in the Roslin Institute – a mix of primary and continuous cells from various anatomical sites: three epithelial cell types (two mammary, one respiratory tract) two immune cell types and four sets of fibroblasts from various organs. Given the representation of different anatomical sites, cell types and differentiation statuses, we considered this a suitably diverse panel with which to characterise infection dynamics of a broad range of IAVs, before more focussed investigations using the bMEC and explant tissues. Both mammary epithelial cell types grew our library of influenza challenge strains significantly better than the BAT-II respiratory epithelial cells, as well as the two immune cell types and all four fibroblast populations. Of the fibroblast cells, those derived from the brain grew IAV significantly better than the skin and turbinate fibroblasts, while blood-derived macrophages grew virus significantly better than the lymphocytes and non-brain fibroblasts. So there are “apple-to-apple” comparisons as well as apple-to-pear comparisons that give significant differences. We therefore think that our conclusions (in the abstract) that mammary cells are particularly replication competent for IAV, (at the end of the introduction) that “a wide range of cow-derived cells are susceptible” and that (in the results section) that “mammary cells showed the highest susceptibility” are justifiable. However, we agree that testing a wider variety of epithelial cells would be useful and have added text to the Discussion (lines 224-228) to acknowledge this.

(Q6) We used a higher MOI for bMECs because test experiments with WT PR8 and the Cattle Texas 6:2 reassortant virus showed that MOI 0.01 infections gave more variable results than those run at MOI 0.1, perhaps because of the intrinsic variability of mixed primary cell populations. However, the end-point titres between the two conditions were not significantly different, so we therefore chose to go with the higher MOI. Accordingly, we do not think this choice is a confounding issue. This explanation (line numbers 340-345) and a new Supplementary Figure 11 showing the results of the two MOI tests have been added to the manuscript.

Reviewer #2 (Public review):

The authors use a library of influenza A viruses from different strains, classified in lab-adapted, human, avian, and swine according to the animal from which they were isolated. They propose that the cow mammary gland serves as a mixing vessel for influenza A viruses. As a first approach, the authors assess susceptibility to infection across different cell types, including continuous and primary cell lines, bovine mammary cells, and mammary explants. All these cells support polymerase activity. Then, they analyzed changes in the bovine virus's viral fitness relative to an avian precursor. The authors use single-gene replacement to study whether and which RNP segments improve viral transcription. As part of this section, they also test IFN-specific antagonism by NS1 to assess the input of segment 8. Quantitative glycomic analysis was performed on the continuous bovine mammary cell line to demonstrate the presence of both a2,3 and a2,6, which is consistent with their observation that these cells can be co-infected with human and avian IAVs simultaneously. The main question, however, is: what is the glycome in the explants, or directly from tissues?

We report quantitative glycomics for the primary bovine mammary epithelial cells as well as the continuous line the referee highlights. However, we agree with R2 that a detailed glycomic analysis of primary bovine mammary tissue would allow a better understanding of the actual glycosylation status in vivo. This has been undertaken by the authors and is available as a bioRxiv preprint. This is now cited (ref 25) in the relevant part of the results (line 184-185)

Overall, the manuscript is clearly written and provides new insights into the behaviour of the cattle isolate, now compared with a representative group of model or precursor HAs of different origins.

It would be great if a consistent nomenclature for the IAV strains could be used in the study. There is a mix of origin (Texas), animal from which the virus was isolated (mallard), or abbreviations that do not follow guidelines (IAV07). Are the USSR and Udorn not lab-adapted?

We chose the abbreviated names for a variety of reasons. Partly from common usage (e.g. PR8, Udorn), partly for consistency with other already published papers from the FluTrailMap consortia (e.g. Cattle Texas; Dholakia et al 2026), partly to make diversity obvious in certain figures (e.g. H3N1, H5N2 etc) and partly to avoid confusion between viruses that originate from the same geographic area (e.g. AIV07, AIV09, H5N8-20 etc which are all A/Ck/England/isolate numbers). Overall, we found it more confusing to use the expanded nomenclature. Re AIV07 which the referee criticises for not following naming guidelines – if this is a reference to the EURL nomenclature, AIV07 is the abbreviation for the specific virus A/Chicken/England/053052/2021, our representative virus for EURL genotype EA-2020-C, as we say in the text. This nomenclature has now been added to Table 1, to provide a fuller cross-reference for all the names.

As to whether USSR and Udorn are lab-adapted – that depends on definitions. There is a continuum of adaptive changes and/or sequence drift starting from the very first growth cycle of an isolate in the laboratory. The viruses we define here as lab adapted are ones that have been deliberately adapted to other host species or which have very long passage histories in multiple laboratory systems resulting in known functionally significant changes; for example, one lineage of PR8 was passaged 77 times in mice, 717 times in cell culture, 30 times in chick embryos, 5 times in ferrets and a further 50 times in chick embryos (https://www.medscape.com/viewarticle/812621_3?form=fpf), rendering it unarguably lab-adapted. We admit that A/USSR/77 and A/Udorn/307/1972 are probably further along this adaptive pathway than more recent isolates such as A/Norway/3433/2018, but are unaware of any specific reason that would put them into our lab-adapted category.

The experimental setup includes bovine mammary primary and continuous cells, as well as mammary explants. Some of the most significant differences, for example, in viral fitness studies and co-infection experiments, are observed in these explants. Perhaps there could be some additional focus on this observation. The implications in comparison to the results obtained in cultured cells could be described. How will the human and other HA subtype viruses fare in the explants?

We agree that this is an important and interesting question, and had already tested the strains we used for co-infections: human seasonal pdm09 H1N1 “Norway” and low pathogenic avian influenza “H3N1”, in the mammary explants. Both replicate the avian virus to 20-fold higher titres. We have added this information to the revised manuscript as new Figure panels S6E-H, called out on line 200-201 of the results.

Reviewer #3 (Public review):

Summary:

This excellent manuscript by Pinto, Sharp, and colleagues examines bovine tissue tropism for influenza viruses. They find that bovine flu, as well as other strains, has strong replication in mammary tissue. They also map the genetic changes to influenza that improve replication in bovine cells. Overall, the study is well designed and executed, and the results are very timely.

Strengths:

(1) The experiments are well-controlled.

(2) The figures are well-constructed and easy to follow.

(3) The Methods and legends are detailed, with sufficient information.

Weaknesses:

(1) A comparison to human cells would strengthen the overall impact of the results. Are human mammary cells also uniquely susceptible to influenza? Are bovine mammary cells special in some way?

This is an interesting question, but we have not tested mammary gland cells from humans (or any other species of mammal). We have however reported elsewhere (Dholakia et al., Nat Commun. 2026 Jan 16;17(1):1603. doi: 10.1038/s41467-026-68306-6.) that Cattle Texas grows well in a variety of human respiratory cells. Here, we are considering the bovine mammary organ as a potential reassortment site for IAVs because of the ongoing viral mastitis epidemic in US dairy cattle; human mammary organs seem unlikely to create a similar opportunity.

(2) For the virus infection studies with segment 8 swaps, it should at least be noted that some of the phenotypes could be driven by NEP.

We agree; as Table S1 indicates, NEP has two changes (one shared with NS1) between AIV07 and our B3.13 isolate, so we should not have conflated segment and NS1. We have changed the text to acknowledge this throughout the results (lines 127, 137 and 149) and in the discussion (line 243-244).

(3) The data demonstrating that bMEC can support co-infection are compelling and important, but would be strengthened with a comparison from a different cell type or species. Do mammary cells uniquely support higher co-infection?

We have data showing that co-infection also occurs in the continuous MAC-T udder cell line and have now included these data in a revised Figure 4D (described/called out on lines 198-206). We have not tested bovine cells from other organs for co-infection potential as they do not seem to be significant sites of infection in vivo.

Recommendations for the authors:

Reviewer #1 (Recommendations for the authors):

(1) How nasal turbinate and cardiac fibroblasts are acquired/cultures is missing from the methods.

Apologies for the omissions and thank you, because rectifying this brought to light an error in cell naming. The cells originally called bovine cardiac fibroblasts were in fact a second independent preparation of skin fibroblasts. We have corrected the labelling in Figs 1, 2 and S2. The nasal turbinate cells were bought in from ATCC (code CRL-1390). This information has now been added to the methods (line 310).

(2) Please specify what cell types make up the 2D enteroids, the mammary explants and the nasal turbinates.

The composition of the 2D enteroids is described in detail in reference [60]. In precis, they are comprised predominantly of epithelial cells, including Paneth, goblet and enteroendocrine cells, as well as stem cells. This information has been added to the methods (lines 374-376) The mammary explants include duct epithelium, connective and muscle tissue, defined by H&E staining of cut sections (see new Figure S12 and text added to the methods on lines 384-386). We have also added the person who did the histology (Rebecca Ross) as an author and to the credit taxonomy (line 877). The nasal turbinate cells appear to be predominantly fibroblast morphology (Methods line 310-311).

(3) Epithelial cells are the main target for IAV, so why were fibroblasts chosen as a comparison to mammary epithelial cells? This needs justification in text. Without justification, how much can be attributed to the results being mammary-specific, rather than epithelial-specific? The brain (choroid plexus epithelium), heart (epicardium) and skin all contain epithelial cells.

We think this query is what the referee calls “Q4’ in the public part of their review. Please see our answer above.

(4) Figure 1, S1, S2 and S3 seem to suggest that mammary cells are more susceptible to IAV infection than cells from other organs. But Figure 2 demonstrates that when it comes to Cattle Texas and AIV07, most of the cell types show high viral titers. If the question is about whether the cow udder is the primary mixing site, would it not be more relevant to investigate which cell type facilitates the best growth of the potential precursor viruses, similar to Figure 4A?

Figs 1, S1, 2 and 3 all use “full” viruses whereas Fig 2 uses 6:2 reassortants between PR8 and the HPAIVs for biosafety reasons. WT PR8 replicates well in most of the bovine cells tested (Figs S1-3) so we do not see any contradiction. Figure 2 examines the contributions the internal genes make to replication in bovine cells. Re the question over the udder being a potential mixing site – this is where the virus is replicating in the real world; at least in part because of the transmission mechanism, but also because the mammary gland epithelium is highly susceptible to infection, as we show.

(5) It is misleading to compare the bMEC infection of MOI 0.1, to all other infections of MOI 0.01. This is a consistent problem throughout the article - Figures 1, S1, 2, 4. Why was the bMEC infection at a 10x greater dose? The main premise of the article relies on bMEC and MAC-T (primary and immortalised mammary epithelial cells), facilitating higher viral growth than the cells from other organs. If we compare the MOI 0.01 experiments alone, then the evidence relies on the immortalised MAC-T cells, compared to primary cell types. In this case, how much can be said about it being mammary specific, rather than immortalised vs primary? I do wonder, for example, how the epithelial nasal turbinates or type II pneumocytes would compare to the primary mammary epithelial cells if they were at the same MOI.

Please see our answer to this query earlier in the rebuttal.

(6) Why are the 2D enteroids excluded from Figure 1?

We had limited supplies of a difficult-to-grow cell model, so we only used them to test the 6:2 viruses (Fig 2).

(7) The colour scheme for Figure 3 is confusing. In Figure 3A, blue indicates European ancestry, and yellow represents North American ancestry. However, in Figure 3B, these colours now mean something different. To a reader, when there is a colour-coded schematic, it is instinctual to think that this then corresponds to the following panel(s). Since consistently throughout the article, yellow has been used for Cattle Texas, and blue has been used for AIV07, I would suggest choosing different colours to represent European and North American ancestry in Figure 3A

We’ve changed the figure as the referee suggests and modified the Fig 3 legend accordingly (line 895)

(8) I am unsure about the conclusions drawn from the results of Figure 3B. In the results, it is framed as trying to determine which segments contributed to the improved activity of Cattle Texas compared to AIV07. In lines 110-111, "... PB2 or PA from AIV07 significantly decreased Cattle Texas minireplicon activity". If PB2 is indeed significant, there is a missing yellow asterisk in Figure 3B.

Apologies, there was indeed a missing asterisk on the figure; now added.

Given the significance of PA, why was it not investigated in terms of growth kinetics similar to Figure 3C? Was it overlooked because it doesn't have a North American ancestry? The results of Figure 3B suggest that the 4 amino acid mutation in PA has significantly contributed to changes in polymerase activity.

The PA changes do indeed matter for minireplicon activity – the key change is K497R, as detailed in our related publication in Nat Comms (citation 17). However, it is less important than changes in PB2, and the PA segment swap by itself has little effect on overall virus replication.

(9) Similarly, in Figure 3C and lines 116-117, the error bars on the graph are overlapping at 48 hours, suggesting no difference in overall replication. The kinetics are slowed for AIV07 seg1-3, but not for AIV07 seg 1, indicating PB1 does not have an effect. This would then suggest that something in segment 2 or 3 contributes to the slowed kinetics in Figure 3C, which, from the Figure 3B results, is unlikely to be due to PB2. While it was not reassorted, the PA segment is potentially the driver, with its 4 amino acid mutations. I think it is worth performing growth kinetics with and without these 4 amino acid changes in PA.

We agree that visually on a log10 scale, the titres of the “WT” 6:2 Cattle Texas and 5:2:1 segment 1 reassortment appear close, but the average titres are 5 and 7-fold different at 24 and 48h respectively, while a 2-way ANOVA with Dunnet’s multiple comparison post-test gives statistical significance at 48h. We have added this information to the figure and its legend (lines 905-906).

(10) In Figure 3C and E, why did you choose to perform the growth kinetics in the immortalised cell line, when you have access to primary cells? The primary cells would be a more accurate representation of what happens in situ.

The primary cells were difficult to work with and only available intermittently, so we used what was available at the time.

(11) In lines 145-147, "thus overall, the reassortment event that replaced segments 1, 2 and 8 alongside drift adaptations in segment 3 may have contributed to the ability of the B3.13 genotype virus to infect cattle". This is not clearly supported by the evidence presented. In terms of segment 1/PB2, the growth kinetics of Figure 3C have overlapping error bars at 48 hours. Where is any evidence presented for the role of segment 2/PB1? There is no change in Figure 3B.

The referee is correct, calling out seg2 here was an error; we have revised the text (line 144).

Segment 3 is overlooked in Figure 3 (as highlighted in Q9 and 10), and shows no difference in Figure S4.

Please see response to Q8; we think segment 3 contributes via PA adaptation, not via PA-X.

(12) In line 146 "... drift adaptation in segment 3". Make it clear here that you are talking about genetic drift. However, is this likely to be genetic drift? The 4 amino acid mutations are shown to have a significant impact on polymerase activity in Figure 3B, and in Figure 3C, PA potentially contributes to the reduced kinetics. When there are amino acid mutations that correspond to a beneficial phenotypic change, attributing this to drift alone rather than host adaptation is strange.

Yes, wording clarified (line 145). “Drift” was used to distinguish it from reassortment but we agree this was an incorrect term in the context.

(13) Figure 4A MAT-C cells: this is ostensibly the same experiment as Figure 3C in terms of the Cattle Texas and AIV07 viruses. If this is the case, how can you explain the difference in kinetics and overall titer? In Figure 3C, Cattle Texas reaches 10^6, and in Figure 4A it reaches almost 10^9. That's almost 3 log difference. Similarly, in Figure 3C Cattle Texas reaches 10^3, but in Figure 4A it reaches 10^6, a 3-log difference. At 24 hrs, they have roughly a 3-log difference between them in Figure 3C, but in Figure 4A this difference is much smaller. As far as I can tell, these are the same viruses, same dose and same cell model. The t0 titer is also vastly different between the two experiments.

The experiments were done at different times (several months apart, so different cell passage numbers and/or serum batches) and by different people. We have no explanation other than biological variability. However, both groups of experiments include genuine biological replicates done over the course of 2-3 weeks, so in our view represent coherent tests within themselves.

(14) In Figure 4, why weren't the a-2,6 a-2,3 proportions analysed for the explants and/or used for the co-infection experiment? It showed the greatest difference between the Cattle Texas and precursor viruses in Figure 4A.

Our data on the proportion of 2,6 and 2,3 SA in bovine udder tissue are now available in a separate preprint (now cited as [25] in our MS). We did not use the explants for co-infection experiments because it would have been technically difficult to read out the outcome by flow cytometry.

(15) Figure 4D requires a supplementary figure demonstrating the gating strategy, including one of the samples as an example.

We have compiled a figure of this and added it as new Figure S13 (called out line 556).

(16) In Figure 4D, why was an MOI of 5 chosen instead of the MOI of 0.01 used throughout the article for MAC-T cell infection? An MOI of 5 (so in a co-infection, a total of 10 virus particles per cell) is completely overloading the cells. At this dose, 10% of the cells were able to be co-infected, but how representative is this of a real co-infection scenario? While it demonstrates it is possible, it potentially remains highly unlikely, similar to the discussion around the swine respiratory tract in lines 235-240.

We had also performed the co-infections at lower MOI (1) with very similar results – this is now included in Figure 4D. Furthermore, we redid the experiments at a lower MOI of 0.05 and still see co-infection; this now replaces the MOI 5 data in Figure 4D. The text has been revised accordingly (lines 202-205)

(17) In Figure 4D, why were immortalised cells used when primary bMEC and mammary explants are available? Primary cells would provide more convincing evidence for the potential of the cow udder to be a mixing vessel. Considering that throughout the paper, a 10x higher viral dose is used in the bMEC culture, I wonder if you would need a significantly higher MOI than 5 to produce similar results in a co-infection experiment. The bMEC also has a more even a-2,3 to a-2,6 ratio compared to MAC-T in Figure 4C.

The bMECs in the original figure are primary cells. In response to other queries, we now include data from the immortalised MAC-T cell line as well.

Reviewer #2 (Recommendations for the authors):

Figure 1A, the coloured underline to discriminate continuous and primary cells is lost upon printing... perhaps another way is better?

We have changed the primary cells to italic text to make the distinction clearer.

We have made some other minor changes to wording to correct grammatical errors or improve clarity as we went through.

Version 2

Abstract

The incursion of high pathogenicity avian influenza A virus (IAV) into US dairy cows is unprecedented in the era of molecular diagnosis and pathogen sequencing. This raises questions over the likelihood of further outbreaks and whether dairy cattle could be a “mixing vessel” for novel strains of IAV. Using a panel of BSL2-safe reassortant viruses representing clade 2.3.4.4b H5 epizootic lineages circulating since 2020, we found that a cow B3.13 isolate displayed enhanced replication in cow mammary gland cells, along with increased viral polymerase activity and stronger interferon antagonism in cow cells compared to an earlier EA-2020-C genotype virus. However, multiple avian and mammalian IAV strains, including other clade 2.3.4.4b high pathogenicity genotypes, were replication competent in bovine cells, particularly those of the mammary gland, suggesting that there is a diverse circulating IAV pool with the potential to infect cows. Moreover, we show that cow mammary cells co-express α-2,3 and α-2,6 – linked sialic acids, and are susceptible to co-infection with human and avian IAVs. We conclude that the US cow influenza outbreak does not simply reflect a unique adaptation of the B3.13 genotype virus; rather, the bovine udder represents a permissive niche for IAV and a plausible site for reassortment, underscoring its potential role in generating novel influenza viruses with pandemic risk.

Introduction

Since 2020, the H5N1 clade 2.3.4.4b of influenza A virus (IAV) has spread globally, threatening diverse wild bird species and causing economic devastation to the poultry industry. This panzootic has also resulted in multiple intra-phylum spill-over events, including farmed mink, semi-aquatic mammals and, in 2024, dairy cows [1, 2]). Since the detection of the B3.13 genotype H5N1 in cattle in March 2024, it has demonstrated sustained, widespread inter-state transmission in dairy cattle [3, 4]. This was unexpected, as until this event, fully verified evidence of IAV infection in cows has been limited, especially with regards to sequenced virus isolates available for further study [5]. While IAV is generally respiratory-tropic in mammals, the primary clinical indicator in the current bovine outbreak is an abrupt drop in milk production[6], correlating with high viral loads in milk[2]. Initial evolutionary and molecular analyses of cattle B3.13 viruses suggested that these viruses had quickly acquired mammalian associated adaptations in several genes [7]. Isolates were replication competent in human cell lines, were highly pathogenic in mice and ferrets [8, 9] and concomitantly, there have been multiple independent reports of zoonotic infections in dairy farm workers [10–12]. These events raised important questions around the likelihood of further IAV incursions into cattle and the pandemic potential of such outbreaks. We show here that a range of cow-derived cells are susceptible to infection with various IAV strains of avian, human or swine origin, that primary cow mammary cells express both α-2,3 and α-2,6 N-acetylneuraminic acid and that these cells can be co-infected with human and avian IAVs. We conclude that other IAV strains have the potential to infect cows and that this could yield reassortant viruses with zoonotic and pandemic potential.

Results

Cow mammary epithelial cells are susceptible to infection with diverse IAV strains

While IAV is traditionally considered to be a respiratory virus in mammalian species, transmission of B3.13 viruses between dairy cows has been associated with the use of milking equipment [13]. We therefore asked whether cow mammary gland epithelial cells are more susceptible to IAV infection than other cell types. Cow primary and immortalised cells originating from different anatomical sites (Figure 1A) were infected with a variety of IAV strains representative of four categories: laboratory-adapted, human, swine and avian. Infections were performed at low multiplicity of infection (MOI) and viral growth kinetic analyses were conducted up to 48 hours post infection (hpi). Of all tested cells, mammary cells showed the highest susceptibility to infection, with lab-adapted strains reaching viral titres of 108 PFU/ml. Titres of 106-7 PFU/ml were detected for viruses of human, swine and avian origins. Respiratory, anatomically diverse fibroblast, and immune cells were less permissive to IAV infection, with peak titres typically 2 to 3 orders of magnitude lower than those achieved in mammary cells across all virus strains (Figure 1B, Supplementary Figures S1-3). This further correlates with the ability of cow mammary cells to support the polymerase activity of diverse IAV strains (Figure 1C). The tropism of B3.13 for mammary gland epithelial cells therefore appears to be a generalisable phenomenon, observed across diverse IAV strains isolated from hosts spanning multiple taxonomic classes.

Cow cells are susceptible to infection with diverse IAV strains.

A). Schematic representation of the anatomical source of immortalised (green, italic text) and primary (black) cells used in this study. B) The indicated cells were infected with a variety of viral strains at MOI 0.1 (bMEC) or MOI 0.01 (for the remaining cells) and virus titres measured by plaque assay at 48 hpi. Viruses were subcategorised into lab-adapted, human, swine and avian strains as indicated. Circles represent individual data points from 3 independently performed experiments, except PBMC (n = 1). Lines represent median. Statistical annotations are the result of a 2-two-way ANOVA analysis. Tukey’s multiple comparisons tests were performed against MAC-T (amber asterisks) or bMEC (burgundy asterisks) data sets. C) RNP-reconstitution assays (minireplicons) were performed in MAC-T cells by transfecting pDUAL plasmids encoding segments 1, 2, 3 and 5 from the indicated viruses alongside a firefly luciferase-expressing vRNA-like reporter plasmid. Luminescence levels were measured at 48h post transfection. Data represent mean ± SEM from three independent experiments performed in triplicate. For statistical annotations: NS denotes non significant and *** a p-value <0.001.

B3.13 internal gene constellation confers better virus replication in cow cells

The inherent permissiveness of cow mammary cells to IAV infection does not explain why B3.13 was particularly successful at sustaining onward transmission in cows. We hypothesised that the unexpected dissemination of B3.13 infections in dairy cows could be explained by a fitness advantage of these viruses over other genotypes. To test this, biosafety level 2 viruses representative of an early B3.13 cow isolate (‘Cattle Texas’) and an avian precursor (AIV07, EURL genotype EA-2020-C) were generated harbouring the glycoproteins of the laboratory-adapted strain A/Puerto Rico/8/1934 H1N1 (PR8). Amino acid differences between the major internal gene products encoded by these viruses are summarised (Table S1). These viruses were used to infect bovine cells from different tissues including mammary epithelial cells (bMEC), type II pneumocytes, skin fibroblasts and blood-derived macrophages (BDΦ) as well as mammary gland explants and surface-grown intestinal organoid cultures (2D enteroids). With the exception of bMECs, where AIV07 replicated slightly better, Cattle Texas replicated to 10 to 10,000-fold higher titres compared to AIV07 (Figure 2). Therefore, Cattle Texas demonstrated a fitness advantage in bovine cells compared to a representative avian virus precursor.

Cattle Texas is fitter in cow cells than an avian EURL genotype EA-2020-C precursor.

The indicated cow cell systems were infected with 2:6 reassortant viruses with PR8 glycoproteins and internal gene constellations of either AIV07 (blue) or Cattle Texas (yellow) (MOI 0.1 in B, 5000 PFU/explant in C and MOI 0.01 for all other systems). Supernatants were collected at the indicated times post infection and virus titres determined by plaque assay. Data represent mean ± SEM from two-four independent experiments each performed in duplicate. Statistical annotations represent the results of multiple unpaired t-tests performed for each individual time post infection. Non significant results are not indicated, * indicates a p-value <0.05, *** = p-value <0.001 and **** a p-value <0.0001.Cell types were immortalised mammary epithelial (MAC-T; A), primary mammary epithelial (bMEC; B), mammary gland explant (C), immortalised pneumocytes (BAT-II; D), skin fibroblast 1 (E), skin fibroblast 2 (F), blood derived macrophage (BDΦ) (G), and 2D enteroids (H).

After entering North America through the Atlantic bird migratory route, EA-2020-C viruses like AIV07 reassorted with the local low pathogenic avian influenza (LPAI) pool, giving rise to the B3.13 genotype via acquisition of new viral genome segments 1, 2, and 5 (encoding the PB2, PB1 and NP components of the viral ribonucleoprotein (RNP) complex respectively), [14] as well as segment 8 (encoding NS1 and NS2) [7, 15, 16] (Figure 3A). Therefore, we investigated whether the new RNP segments improved viral transcription activity in cow cells using a bovine-specific minigenome reporter assay [17]. Cattle Texas RNPs were 20-times more efficient than AIV07 (Figure 3B, open blue and yellow bars). To determine which segments contributed to this improved activity, Cattle Texas and AIV07 RNPs were reconstituted with reciprocal single gene replacements, including segment 3 (encoding PA), the final component of the viral polymerase. Replacing Cattle Texas PB1 with the AIV07 counterpart had no effect, while swapping NP increased RNP activity in MAC-T cells (Figure 3B, yellow cross-hatched bars). In contrast, PB2 or PA from AIV07 significantly decreased Cattle Texas minireplicon activity. In the reciprocal swaps, AIV07 polymerase activity was improved by up to 5-fold by Cattle Texas segments 1, 2 or 3, but replacement of segment 5 had an 8-fold negative impact on polymerase efficiency (Figure 3B, green cross-hatched bars). Thus, the increased polymerase activity of Cattle Texas is attributable to the PB2- and PA-encoding segments. To test this in a viral context, we made reassortant viruses in which Cattle Texas PB2 or all polymerase segments were replaced with those of AIV07. Both PB2 and full polymerase swap viruses exhibited reduced replication in cow mammary cells (Figure 3C), confirming the importance of these segments.

Increased fitness of Cattle Texas in cow cells correlates with enhanced viral polymerase activity and better NS1-driven inhibition of type I IFN induction.

(A) Schematic representation of AIV07 and Cattle Texas segment origins. Blue and green represent European and North American ancestry, respectively. (B) RNP-reconstitution assays (minireplicons) were performed in MAC-T cells by transfecting pDUAL plasmids encoding segments 1, 2, 3 and 5 from AIV07 and Cattle Texas in the indicated combinations alongside a firefly luciferase-expressing vRNA-like reporter plasmid. Luminescence levels were measured at 48h post transfection. Data represent mean ± SEM from three independent experiments each performed in triplicate. Statistical annotations are the result of one-way ANOVA analysis. Multiple comparisons were made against full AIV07 3PNP (blue asterisks) or Cattle Texas 3PNP (yellow asterisks) using a Dunnett’s test (C) MAC-T cells were infected with polymerase-segment reassortment viruses at MOI 0.01. Supernatants were collected at the indicated time points and virus titres measured by plaque assay. Data represent mean ± SEM from three independent experiments each performed in technical duplicate. Statistical annotations are for 48h titres following a two-way ANOVA analysis and multiple comparisons using Dunnet’s test made against WT 6:2 Cattle Texas. (D) MAC-T cells were transfected with an IFN-β firefly luciferase-coding reporter alongside AIV07 or Cattle Texas segment 8 plasmids, and poly I:C treated 24h later. Luminescence readings were acquired after a further 24h. Data (left hand panel) represent mean ± SEM from three independent experiments each performed in technical triplicate. Statistical annotations are the result of one-way ANOVA analysis followed by a multiple comparison Dunnett’s test. A western blot (right hand panel) was also performed to detect NS1, using α-Tubulin as a loading control. The positions of molecular weight markers are indicated. (E) MAC-T cells were infected with segment 8 reassortant viruses at MOI 0.01. Supernatants were collected at the indicated time points and virus titres measured by plaque assay. For all shown statistical annotations: NS indicates non significant, * a p-value <0.05, ** a p-value <0.01, *** a p-value <0.001 and **** a p-value <0.0001.

NS1 is the primary IAV IFN antagonist, working through multiple partially redundant but also strain-specific mechanisms, including both general repression of cellular gene expression at transcriptional and translational levels as well as specific inhibition of IFN induction [18]. As an initial test of whether reassortment of segment 8 was phenotypically important, we assessed host cell shut off by infecting MAC-T cells with the PR8 6:2 Cattle Texas or a PR8 2:5:1 Cattle Texas harbouring AIV07 segment 8 and pulsed the cells with puromycin at various times post infection. Western blotting for the puromycin-labelled prematurely terminated polypeptides indicated that while both viruses repressed cellular translation at late times post infection, there was no major difference between them (Figure S4A). As a more quantitative measure of shut off, we co-transfected MAC-T cells with segment 8 plasmids from the two viruses alongside a constitutive expression plasmid encoding Renilla luciferase and measured the resulting luminescence. Both Cattle Texas and AIV07 segment 8 gave dose-dependent inhibition in this assay with Cattle Texas showing a modest improvement over AIV07 (Figure S4B). The PA-X accessory gene expressed from segment 3 also contributes to IAV host cell shut off [14]. Although segment 3 reassortment did not occur between genotypes EA-2020-C and B3.13, genetic drift led to four amino acid differences between the PA-Xs of AIV07 and Cattle Texas, three of which lie within the “X-ORF” (Figure S4C). Therefore, we similarly compared the ability of AIV07 and Cattle Texas PA-X proteins to inhibit host gene expression, but no significant differences were observed (Figure S4D).

To test specific IFN antagonism by NS1, the AIV07 and Cattle Texas segment 8 plasmids were transfected into MAC-T cells alongside an IFN-β reporter plasmid and the cells stimulated with poly I:C as a mimic for viral RNA. This showed that, at similar expression levels, Cattle Texas segment 8 had a significantly higher ability to counteract type I IFN induction than that of AIV07 (Figure 3D). To assess the contribution of this to overall virus fitness, replication of the 2:5:1 Cattle Texas reassortant virus harbouring AIV07 segment 8 was characterised in MAC-T and BAT-II cells. Introduction of AIV07 NS into Cattle Texas resulted in a slight but non-statistically significant delay in virus propagation in mammary epithelial cells, but had a much larger effect in the pneumocyte cell line (Figure 3E), suggesting that changes in segment 8 also contribute to the improved fitness of Cattle Texas over AIV07. Thus overall, the reassortment events that replaced segments 1 and 8 alongside genetic adaptations in segment 3 may have contributed to the ability of the B3.13 genotype virus to infect cattle.

Spill-back of the cattle B3.13 virus into poultry has been observed on multiple occasions [3, 19], despite its evident fitness adaptations for bovines. The 2:6 Cattle Texas virus replicated to slightly lower titres in chicken lung epithelial cells than AIV07, while its polymerase and NS1/NS2 exhibited similar activities to those of the AIV07 proteins in the same cells (Supplementary Figure S5). Thus, the enhanced fitness of Cattle Texas in bovine cells is specific to bovine cells but does not carry a high fitness penalty in chicken cells.

Cow mammary cells are susceptible to European 2.3.4.4b genotype viruses and support co-infection by human and avian viruses

Detections of IAV in US dairy cattle are mostly of North American H5N1 2.3.4.4b genotypes B3.13 and, more recently, D1.1 viruses. Therefore, we asked whether similar dairy cattle incursions could happen in countries where the B3.13 and D1.1 genotypes are not circulating. To that end, we compared the growth kinetics of Cattle Texas with other avian European 2.3.4.4b genotypes in primary and continuous bovine mammary cells and explants, again all as 2:6 reassortants with PR8 glycoproteins to permit work at BSL2. In addition to AIV07 (genotype EA-2020-C), these included a 2.3.4.4b 2020 H5N8 precursor (H5N8-20) and EURL genotypes EA-2021-AB (AIV09) and EA-2022-BB (AIV48) (Figure 4A). In all three systems, Cattle Texas replicated to the highest titres, perhaps as expected for a virus with bovine-adapted internal segments. Other unadapted 2.3.4.4b strains showed varied viral fitness (Figure 4A). Cow mammary explants showed the greatest range of virus titres, with H5N8-20 and AIV07 exhibiting very poor fitness in this model, such that many of the explants were refractory to infection or only yielded low titres (Figure S6A-D). However, the more recent AIV09 and AIV48 isolates replicated better in the explants, and Cattle Texas demonstrated the highest virus fitness, consistent with a trend for increasing replication fitness in mammary tissue of the 2.3.4.4b strain over time. This suggests that recent Clade 2.3.4.4b H5N1 genotypes on both sides of the North Atlantic have the potential to cause dairy cattle infections.

Cow mammary cells are infectable with viruses of human and avian origin.

(A) MAC-T (MOI 0.01), bMEC (MOI 0.1) cells and mammary explants (5000 PFU/explant) were infected with the indicated PR8 2:6 viruses. Supernatants were collected at the indicated times post infection and virus titres determined by plaque assay. Statistical analysis was performed by a 2-way ANOVA and multiple comparisons were performed against Cattle Texas in specific time points using a Dunnett’s test. (B) Annotated MALDI-TOF spectra of permethylated N-glycans from MAC-T and bMEC cells. Annotations show [M + Na]+ molecular ions. Peak annotation is based on composition, biosynthetic knowledge and MS/MS analysis. (C) Proportions of a-2,3 and a-2,6 linked sialic acid in MAC-T and bMEC N-glycans. (D) MAC-T and bMECs were infected with combinations of Norway (pH1N1) and/or Belgium (H3N1) at MOI 0.05 for 48h or MOI 1 for 16h. Infections at MOI 0.05 were overlayed with media containing TPCK-treated trypsin. Cells were fixed, stained with anti-H1 and H3 before quantification was performed by FACS analysis. Data represent mean ± SEM from three independent experiments each performed in technical duplicate. For all shown statistical annotations: ****p-value <0.0001.

Human IAV pandemics classically arise from reassortment between avian and mammalian strains [20, 21]. Historical data suggest the incursion of human seasonal strains into cattle [5], so in the face of an ongoing H5N1 epidemic in birds and cattle, the possibility of the cow udder as a “mixing vessel” arises. However, co-infection of a single cell harbouring both avian (α-2,3 linked sialic acid) and mammalian (α-2,6 linked sialic acid) receptor types is needed for this mixing to occur. Lectin staining and mass spectrometry analyses indicate the presence of both α-2,3 and α-2,6-linked sialic acid receptors in cow mammary glands [22–25]. To establish the amounts of each receptor type in the cells used here, we next performed quantitative glycomic analyses on the immortalised and therefore likely relatively homogeneous (MAC-T) and primary, potentially heterogeneous (bMEC) cell populations. This identified complex N-glycans containing a mix of α-2,3 and α-2,6 linked sialic acids, with a higher proportion of α-2,6 linkages present on the bMECs compared to MAC-T cells (Figure 4B, C; Supplementary Figures S7-S10). Cow mammary cells therefore express the sialic acids required for infection with both avian and mammalian IAVs; however, this does not demonstrate that the two sialic acid conformations are present at functionally relevant levels on the same cell. To determine whether individual cow mammary gland cells could be co-infected by human and avian strains of virus, MAC-T and bMEC cells were inoculated with an α-2,6 sialic acid-dependent human seasonal H1N1 strain and an α-2,3 sialic acid-dependent low pathogenicity avian virus (both of which replicate in mammary explants (Figure S6E-H), the most discriminatory bovine cell system for replication of the 6:2 reassortant HPAIVs (Fig 4A)), either singly or together and at low or moderate MOI. Infected cells were fixed and stained with anti-H1 or H3 HA and analysed by flow cytometry. The individual infections confirmed antibody specificity, while dual-infected cell populations were clearly identified in all co-infection samples (Figure 4D). We therefore conclude that co-infections of human and avian viruses within the cow mammary glands are possible, supporting the possibility of reassortment events.

Discussion

The unexpected influenza outbreak in US dairy cows requires careful consideration to understand why it occurred, and its potential implications for both livestock and human health [15]. At one extreme, it could have resulted from a rare adaptation of the B3.13 genotype to allow infection of a species usually refractory to infection with most strains of IAV; at the other, an unusually large and well documented epidemic of a not uncommon spillover event that mostly goes undetected. To experimentally test these possibilities, we compiled a variety of bovine cell culture systems and performed a systematic analysis of their susceptibility to IAV infection. We found that while there are clear virus strain and cell-type differences in the susceptibility of bovine cells to IAV infection, many cell types are susceptible to infection with a diverse range of avian and mammal viruses. Notably however, we think our data support mammary gland epithelial cells to be the primary site of infection compared to respiratory, immune or fibroblast cells, consistent with the clinical manifestations of mastitis and high viral loads present in milk [4, 26]. However, the lack of a wide selection of epithelial cell types from other anatomical sites in the cow means that although we can be confident that bovine mammary cells are highly permissive to IAV infection, we cannot rule out that other organs might be equally or even more susceptible if epidemic transmission were to occur by another route than milking equipment. The bovine cells used in our study come from a mixture of dairy and beef cattle. The mammary explants were derived from dairy cattle of unspecified breeds, while MAC-T cells are from Holstein Friesian dairy cattle. The bMECs were derived from beef cattle, and the enteroids from male Holstein Friesian calves. We can therefore be confident that in vitro cell permissivity to IAV infection is not unique to a single breed, sex or age of cattle, but we cannot rule out that these factors might influence viral susceptibility in vivo [27].

The ability of cow mammary cells to be infected with and support high levels of replication of a wide range of IAV strains here described argues against the extreme hypothesis of a truly unique phenotypic trait of the B3.13 strain that permits infection of cows. These observations are consistent with recent studies showing that other avian 2.3.4.4b isolates are able to replicate in bovine cells [28–30] and udders [26, 31], as well as the more recently detected D1.1 genotype spillover into US dairy cattle [32]. However, our findings of wide variation in the ability of specific strains of IAV to replicate in bovine cells suggests that adaptation does play a role. Thus, we sought to determine what, if anything, predicted the emergence and transmission of Cattle Texas above other avian IAV strains. We found a multifactorial scenario, with improvements in polymerase efficiency and the ability of segment 8-encoded products (most likely NS1) to counteract type I interferon induction. These improvements were apparent in cow but not chicken cells, indicative of evolutionary adaptations to the new host. This is consistent with other studies showing a replicative advantage of cattle isolates over an avian precursor [8, 31]. For PB2 and PA, the amino acid mutations PB2 M631L and PA K497R present in the Cattle Texas strain we used have recently been reported to enhance polymerase efficiency and virus replication in cow mammary cells [3, 17, 19]. Furthermore, impairment to transcriptional efficiencies rendered by the newly acquired segment 5 imply that there is still the potential for further fitness gains of the B3.13 genotype in cattle. Notably, the D1.1 strain isolated from cows in Nevada possess PB2 D701N [33] while a human isolate contained PB2 E627K [34], suggesting that other adaptation pathways in cattle are possible [35, 36].

Pigs are the classic reassortment host of IAV, the respiratory tract of which is unusual in its ability to support infection with both avian and mammalian IAV strains and from which the most recent influenza pandemic in 2009 arose [37, 38]. However, HPAI H5 infections in pigs are rare; pigs were either not susceptible to H5N1 2.3.4.4b strains following nasal exposure [39], or when infected no transmission was observed [40]. Therefore, reassortment of H5 HPAI with other mammalian IAV strains within the swine respiratory tract can be considered unlikely. However, our observations on the ability of avian and mammalian IAV strains to infect cow mammary glands raised the question of whether cow udders could serve as tissues permissive for viral reassortment for the generation of novel strains containing human-adapted and H5 HPAI-origin segments, analogous to the pig reassortment events that are conjectured to have precipitated previous pandemics. The fact that primary cow mammary cells can be co-infected with human and avian IAV strains, supported by the mixed distribution of α-2,3 and α-2,6 linked sialic acid observed in these cells, suggests that cow udders do have this potential. However, our experiments necessitated culture of these cells and culture passage has been observed to increase overall sialylation in some cell types [41, 42], meaning that our results may modestly overestimate the total sialylation abundance.

Infection of cows with human and other IAV strains has been suggested in the past. In 1997, dairy cows in the southwest of England presented with sporadic drops in milk production, which the reporting authors considered to be attributable to IAV [43, 44]. Additional studies have reported seroconversion against circulating seasonal H3N2 virus in British cattle, which was associated with reduced milk yield and respiratory disease [45, 46]. Despite the occasional respiratory signs, such as nasal discharge and increased respiratory rate followed by seroconversion, virus isolation from respiratory swabs from these seroconverted animals was unsuccessful. With hindsight, isolation from milk may have had a higher chance of success. Consequently, the idea that “cows could foster flu pandemics” was considered back in the early 2000s [47]. Moreover, the recent seroconversion for H5N1 2.3.4.4b in sheep and detections of IAV RNA in milk [48, 49] further suggests that other ruminant livestock species may hold similar potential. Until recently, the absence of H5 detection in cattle beyond the United States might have indicated that the coordinated polymerase and NS1 and/or NS2 adaptations that we have linked to this type of spillover have not yet emerged in Europe or other regions. It could also be attributable to insufficient surveillance, potentially due to the unexpected presentation as mastitis. However, H5 seropositive dairy cattle in the Netherlands have recently been reported [50]. At the time of writing, no viral sequences have been made available, so whether this represents similar viral adaptation to what we describe for the B3.13 genotype and/or enhanced surveillance remains to be determined.

Overall, our findings point to requirements for reconfigured and enhanced policy procedures and surveillance of IAV incursions into ruminant livestock species. Currently, mastitis outbreaks are not routinely investigated for the presence of IAV. Adding IAV testing to mastitis diagnostic assessments as a differential when other common causes have been excluded will be beneficial to livestock. This will allow early detection of emerging events with implementation of appropriate infection control measures and justify contact tracing of farm workers.

Materials and Methods

Ethics

All reverse genetics work used a loss-of-function approach and was carried out under a license (GMRA1811) from the UK Health & Safety Executive. All work with avian viruses was carried out in a segregated BSL2 laboratory in the Roslin Institute, separate from where mammalian IAV strains are handled.

All animal experimental work was completed under license and in accordance with the Animal (Scientific Procedures) Act 1986 and with approval from the Roslin Institute animal welfare and ethics committee under the project licence number PP0612543. All studies were performed in line with the Animal Research: Reporting of In Vivo Experiments (ARRIVE) Guidelines [51].

Cell line maintenance, primary cell isolation, derivation and culture

Madin-Darby Canine Kidney (MDCK) (American Type Culture Collection, ATCC), MDCK expressing galline ANP32A (MDCK-GgANP32A) [52], MDCK expressing Sialyltransferase 1 (MDCK-SIAT) [53], human embryonic kidney (HEK293T), immortalised mammary epithelial cells (MAC-T) [54] and bovine nasal turbinate (BT) cells (ATCC, CRL-1390; predominantly fibroblast morphology) were cultured in DMEM supplemented with 10% foetal bovine serum (FBS), 2mM glutamine, 100 U/ml penicillin, and 100 μg/ml streptomycin. MDCK-SIAT cells were further supplemented with 50 µg/ml geneticin. MAC-T cells were further supplemented with 5 µg/ml insulin and 1 µg/ml hydrocortisone. Chicken lung cells (CLEC213) [55] were cultured with DMEM/Nutrient Mixture F-12 also supplemented with 8% FBS, 100 U/ml penicillin and 100μg/ml streptomycin. BAT-II cells [56, 57] were maintained in Small Airway Growth medium (Lonza) further supplemented with 10% FBS, 100 U/ml penicillin and 100μg/ml streptomycin. All immortalised cells were regularly tested for mycoplasma contamination and passaged once or twice weekly.

Primary cow mammary epithelial cell (bMEC) lines were generated from the mammary glands of Aberdeen Angus cross cows as previously described [58]. Glands were sprayed with 70% ethanol and skin was removed. Cuboidal epithelial tissue fragments (5 cm in length) were cut from the gland, submerged in 70% ethanol for 30s and washed 5 times in PBS containing antibiotics and 2.5 μg/ml amphotericin B. Tissue fragments were transferred to a glass petri dish and minced with scalpels to 2-3 mm pieces. Next, 10-15g of minced tissue was dissociated overnight with low shaking in 20 ml of DMEM/Nutrient Mixture F-12 supplemented with 15 mM HEPES and 0.075% sodium bicarbonate, 2% Bovine serum albumin (BSA), 300 U/ml Collagenase Type I and 100 U/ml Hyaluronidase from bovine testes. Dissociation media was removed by centrifugation and pellet was washed twice with DMEM/F12. This yielded small aggregates of part-digested tissue which were resuspended in freezing medium and stored at −150°C. To obtain a single cell suspension, aliquots of part-digested tissue aggregated were thawed, washed with Hanks’ Balanced Salt Solution (HBSS) supplemented with 2% FBS, resuspended in warm trypsin and vigorously pipetted for 3 mins and washed with HBSS/2% FBS, resuspended in a warm Dispase/DNAse solution and gently pipetted for 3 minutes. Ice cold HBSS/ 2% FBS was added and the suspension was filtered on a 70 μM strainer, cells were washed by centrifugation, resuspended in growth media (DMEM/F12 supplemented with 0.1% BSA, 10 ng/ml epidermal growth factor (EGF), 10 ng/ml cholera toxin, 1 µg/ml bovine insulin, 0.5 µg/ml hydrocortisone, 100 U/ml penicillin, 100 μg/ml streptomycin and 250 µg/ml gentamycin) and seeded onto collagen coated wells/flasks. bMECs were examined everyday by light microscopy to check for signs of contamination, viability, morphology, and confluence. Media was refreshed daily. Growth media was further supplemented with 2.5 µg/ml amphotericin B, the concentration of which was halved every day until the cells had reached sufficient confluency to be expanded. Initial tests of the permissivity of bMECs to infection with IAV (PR8 and the Cattle Texas 6:2 reassortant) indicated that 0.01 MOI infections gave more variable outcomes than infections at an MOI of 0.1 (Figure S11), perhaps because of heterogeneity in the cell populations. The average end point (48h) titres were not significantly different between the two MOIs, so for further experiments an MOI of 0.1 was used to increase the reproducibility of the data.

Primary fibroblasts were extracted from skin (two independent preparations; referred to as ‘skin 1’ and ‘skin 2’) and brain tissue explants dissociated with 5 mg/ml Collagenase type I at 37°C, shaking for 3h. The dissociated material was passed through a 70 µm filter and washed with HBSS. Cells were cultured on bovine gelatine-coated tissue culture plastic and expanded in Glasgow’s MEM) supplemented with 10% foetal bovine serum, 2 mM L-glutamine, 1x MEM non-essential amino acids, 1 mM sodium pyruvate, 0.1 mM 2-mercaptoethanol, 100 U/ml penicillin-streptomycin, 50 µg/ml gentamicin and 2.5 µg/ml amphotericin B. Established cultures were maintained in the same medium without antibiotics.

Cow PBMCs were collected from Holstein-Friesian cattle housed at the University of Edinburgh farm using published procedures [59]. Cow blood derived macrophages (BDΦ) were generated from PBMCs as described previously [59] except that the macrophages were cultured in BDΦ medium (RPMI-1640 supplemented with 20% FBS, 4 mM L-glutamine and 50 µM β-mercaptoethanol, 100 U/ml penicillin-streptomycin and 20 ng/ml His-tagged recombinant bovine CSF1 - the latter supplied by the Roslin Immunological Toolbox). BDΦ medium was replaced on day 7. On day 12 the adherent cells were rigorously washed with phosphate-buffered saline PBS and detached with TrypLE Express. Purified BDΦs were resuspended at 1.5 x 105 cells/ml in BDΦ medium without CSF1, aliquoted into 24 well plates (1ml/well) and cultured for 48h before infection on day 14.

Bovine 3D enteroids were derived from five 6-month-old healthy male British Holstein–Friesian calves housed at the University of Edinburgh farm. Calves were subject to Schedule 1 method of cull and sections of distal ileum were obtained post-mortem and stored in HBSS on ice until further use. Crypts were isolated and 3D enteroids were cultured as previously described [60]. Bovine 3D organoid cultures were used to generate 2D enteroids as previously described [61]. The 2D enteroids are comprised of predominantly epithelial cells including Paneth, goblet and enteroendocrine cells, as well as stem cells.

Bovine mammary gland explants were extracted from udder tissue collected at Sandyford Abattoir (Paisley, Scotland) from healthy cows of unspecified breeds that had been culled for the food chain. Excess tissue was removed, teat and gland cistern were dissected (Figure S12A) and transferred to transport media comprising of chilled DMEM supplemented with 100 μg/ml Penicillin/ 100 μg/ml streptomycin and 5 μg/ml fungizone. Every 45 minutes, tissues were transferred to 500 ml of fresh transport media. Inside class II biological safety cabinet, explants were made using a 5 mm biopsy punch. Biopsies were added to 24 well plates with 500ul of transport media further supplemented with 10% FBS and incubated for 24h at 37°C, 5% CO2 prior to infection. Histological examination of the explants showed that they consist of duct epithelium over successive layers of connective and muscle tissue (Figure S12B).

Validation of bovine origin of the primary and continuous cell lines was performed by PCR. DNA was extracted from cell pellets, and the cytochrome B gene was amplified by PCR with degenerate primers (CGAGACGTVAAYTAYGGMTGA and ACBGTDGCYCCTCARAADGATATTTG). Cycling conditions were 94°C for 2 minutes, followed by 50 cycles of 94°C for 18 seconds, 50°C for 21 seconds then 72°C for 30 seconds with a final extension of 72°C for 5 minutes. Amplicons were sent for Sanger sequencing and resultant sequences were aligned to the Bos taurus mitochondrion complete genome reference sequence (NCBI accession number NC_006853.1).

All cells, organoids and explants were kept at 37°C and 5% CO2 under normoxic conditions.

Viruses

Selected viruses were split into 4 different categories based on their history and isolation animal host (Table 1). A/Puerto Rico/8/1934 H1N1 (“PR8”) [62], mouse adapted A/California/04/2009 H1N1 (“Cal04”) [63], A/WSN/33 H1N1 (“WSN”), swine adapted A/mallard/Netherlands/10-Cam/1999 H1N1 (“Mallard”) [64, 65], A/Udorn/307/1972 H3N2 (“Udorn”), A/England/195/2009 pH1N1 (“Eng195”), A/swine/England/87842/1990 H3N2 (“Swine H3N2”) and 2:6 reassortants of 2.3.4.4b H5N1 viruses containing PR8 segments 4 and 6 viruses were generated by reverse genetics as previously described[66]. Briefly, 2 × 106 HEK293T cells were transfected in OptiMEM with pDUAL reverse genetics plasmids (250 ng of plasmid for each virus segment), using 4 µl of Lipofectamine 2000 according to the manufacturer’s instructions. At 24 hours post transfection, the media was replaced by serum-free DMEM supplemented with 0.14% (w/v) fraction V BSA and 1 μg/ml of l-(tosylamido-2-phenyl) ethyl chloromethyl ketone (TPCK)-treated trypsin. Virus-containing supernatant was collected after a further 2-days incubation (here described as a “P0 stock”). P0 stocks of PR8, Swine H3N2 and Udorn were propagated in MDCK cells. P0 stocks of Mallard, Eng195 and PR8 2:6 2.3.4.4b H5N1 reassortants were further propagated in 10-day embryonated hen’s eggs. Allantoic fluids were collected 48hpi. Cal04 P0 stocks were propagated in MDCK-SIAT cells. Reassortants between A/dairy cattle/Texas/24-008749-001-original/2024 H5N1 (“Cattle Texas”) and A/chicken/England/053052/2021 H5N1 (“AIV07”) were propagated in MDCK-GgANP32A.

Isolates of A/USSR/90/1977 H1N1 (“USSR”) A/Brisbane/10/2007 H3N2 (“Brisbane”), A/Norway/3433/ 2018 H1N1 (“Norway”) and A/swine/453/England/2006 H1N1 (“Swine H1N1”) isolates were propagated in MDCK cells. Clarified supernatants were typically collected at 36–72 hpi. Isolates of A/chicken/Belgium/460/2019 H3N1 (“H3N1”), A/chicken Belgium/150/1999 H5N2 (“H5N2”), A/chicken/Netherlands/917/2010 H6N1 (“H6N1”) and A/chicken/Saudi Arabia/2525/2000 H9N2 (“H9N2”) isolates were propagated in hen’s eggs and allantoic fluids were collected and clarified at 48 hpi. All viruses were sequenced to confirm identity using Nanopore sequencing (described below).

RNA extraction and Multi-segment RT-PCR

To confirm the genome sequences of the virus strains used, viral RNA was extracted from cell- and egg-grown viral stocks using the QIAamp Viral extraction mini kit (QIAGEN) as per manufacturer’s instructions. Viral genomes were amplified using the Superscript III One-Step RT-PCR kit (Invitrogen), 5 µl of extracted viral RNA, GGGGGGAGCAAAAGCAGG and GGGGGGAGCGAAAGCAGG as forward and GGGGTTATTAGTAGAAACAAGG as reverse primers[67]. Thermocycling conditions for RT-PCR reaction were: 55°C for 2 mins, 42°C for 60 mins, 94°C for 2 mins; 5 cycles of 94°C for 30 s, 44°C for 30 s and 68°C for 3 mins 30 s; 26 cycles of 94°C for 30 s, 57°C for 30 s and 68°C for 3 mins 30 s and 10 mins hold at 68°C. Resulting PCR was analysed on an 1% agarose gel.

Nanopore sequencing and data analysis

RT-PCR products were pooled and prepared for sequencing using a ligation sequencing amplicon kit (SQK-NBD114.24, Oxford Nanopore Technologies, Oxford, UK) and sequenced on a FLO-MIN114 flow cell (R10.4.1., Oxford Nanopore Technologies) on an in-house GridION device. Final library concentration was approximately 10 fmol. Molarity was determined using Oxford Nanopore recommended tables, assuming an average multi-segment RT-PCR fragment size of 1.3 kb. Reads were called using the high accuracy basecall setting on the MinKNOW UI (Oxford Nanopore Technologies). The “passed” FASTQ output files were used in the sequencing analysis. Data was analysed using the EPI2ME (Oxford Nanopore Technologies) and Sequeduct (Edinburgh Genome Foundry) [68] pipelines. The processed FASTQ files were aligned against expected sequences. For the Sequeduct pipeline, reference sequences were converted to Genbank format. Consensus sequence output from the “analysis” workflow was used to align against expected sequences.

Plaque assays

Confluent monolayers of MDCK cells (1 × 106 cells seeded in 12-well plates the day prior to infection) were washed once with PBS, then infected with 250 µl of tenfold serial dilutions of virus. Following a 1 h incubation at 37 °C that allowed virus adsorption, cells were overlaid with DMEM supplemented with 0.14% fraction V BSA, 1 µg/ml TPCK-treated trypsin and 1.2% Avicel. Following a 48h undisturbed incubation at 37 °C, cells were fixed with 10% Neutral Buffered Formalin for at least 20 minutes and stained with a 0.1% toluidine blue solution for at least 1 h and plaques were then counted. Plaque assays for avian viruses or PR8 2:6 viruses with avian origin PB2 were performed as described but using MDCK-GgANP32A cells. Plaque assays for human viruses Eng195, Cal04, Brisbane and Norway H1N1 were performed using MDCK-SIAT cells and incubation temperatures of 35 °C for 72h.

Virus replication kinetics

Monolayers of cells were washed once with PBS and infected with virus diluted in serum-free medium for 1 h at 37 °C (the multiplicity of infections used is specified in each figure legend). Infectious medium was replaced with serum-free medium supplemented with 0.14% fraction V BSA and 1 µg/ml TPCK-treated bovine pancreas trypsin.

Individual mammary explants were transferred to 96-well plates and infected with 5000 PFU of virus in 100 µl of serum free DMEM for 2h at 37°C. Explants were transferred to clean 24-well plates, washed 3x with warm PBS and overlayed with 1ml of DMEM supplemented with 0.14% fraction V BSA and 1µg/ml TPCK. Supernatant was collected at different timepoints post infection and infectious titres were determined by plaque assay.

Immunoblotting

Cells were lysed in Laemmli buffer, heated at 80 °C for 5 min and subjected to polyacrylamide gel electrophoresis using sodium dodecyl sulfate-based buffer. Peptides were transferred to polyvinylidene difluoride membranes for 90 minutes at 32 V. Membranes were blocked for 1 h with 1× Tris-buffered saline with 0.1% Tween20 (TBST)/5% milk, washed three times with TBST and stained with primary antibodies diluted in TBST/5% BSA overnight at 4 °C. After three 5-minute TBST washes, secondary antibodies were diluted in blocking buffer and incubated for 1 h at room temperature protected from direct light. Following three more washes in TBST, membranes were imaged using the LI-COR Fc-Odyssey Imaging platform using the Image Studio Lite Software.

Plasmid-based IFN- β promoter reporter assays

Under confluent monolayers of MAC-T or CLEC213 cells were co-transfected with 500ng or 100 ng of reporter plasmid encoding firefly luciferase under the control of human [69] or chicken IFN2 [70] promoter and 1000ng of segment 8 pDUAL plasmids using 2 μl of Lipofectamine 2000. 24 hours post transfection cells were stimulated with 5 μg of Poly I:C by transfection using 2 μl of Lipofectamine/well. Mock-stimulated cells were also included where Poly I:C was replaced by water. 24 hours post-stimulation, cells were lysed with 100 μl of reporter lysis buffer and luciferase activity was measured from 40 μl of lysate in opaque 96-well plates by using 25 μl of luciferase assay reagent in Cytation3 Plate Reader and Imager (integration time 1 second).

Cellular gene expression shutoff transfection-based assays

Sub-confluent monolayers of MAC-T and CLEC213 cells were co-transfected in triplicate with a range of pDUAL segment 3 or 8 effector plasmids and 200ng of pRL using 2μL of Lipofectamine2000. Forty-eight hours post transfection, cells were lysed and Renilla luciferase expression was quantified using a Renilla Luciferase Assay kit. Briefly, cells were lysed with 100μL of 1× lysis buffer and frozen overnight at −20°C. Cells were scraped off the plate, cell debris was removed by centrifugation (8000rpm, 5 minutes, 4°C). Luminescence was measured from the mix of 20 μl of lysate and 50 μl of 1× Renilla luciferase substrate using a Cytation3 Plate Reader and Imager (integration time 1 second).

Puromycin-labelling of nascent proteins

Subconfluent monolayers of MAC-T cells were infected at MOI 5. At 9, 12, 15 and 18 hpi, supernatant was replaced with fresh medium supplemented with 20 μg/ml of puromycin dihydrochloride for 1h. Cells were washed twice with PBS, lysed in Laemmli buffer and subjected to immunoblotting using a primary anti-puromycin antibody (Millipore; MABE343).

RNP reconstitution (minireplicon) assays

Sub-confluent monolayers of MAC-T or CLEC213 cells (1 × 105 cells seeded in 24-well plates the previous day) were co-transfected in triplicate with 50 ng of each pDUAL plasmid encoding PB2, PB1, PA and NP along with 20 ng of a bovine PolI-driven vRNA-like firefly luciferase reporter plasmid [17]. As negative controls, transfections lacking the PB2 plasmid (with empty pDUAL vector used to balance plasmid amount) were also performed. Two days post transfection, cells were lysed with 120 µl of reporter lysis buffer (Promega E4030). Lysates were cleared and luminescence was measured from 40 μl of lysate in opaque 96-well plates by using 25 μl of luciferase assay reagent in a Cytation3 Plate Reader and Imager (integration time 1 second).

Tissue homogenisation

Mammary explants were added to 1ml of serum-free DMEM and a 5mm diameter stainless steel bead in 2ml tubes and homogenised using a Tissue Lyser II (Qiagen) system at 28kHz for 3 sets of 2 minutes. Homogenised tissue was clarified by centrifugation (3000 rpm, 5 min, 4°C). Homogenates were titrated for infectious virus by plaque assay.

Glycomic analysis of MAC-T and bMEC cells

Approximately 10 million cultured MAC-T and bMEC cells were washed and scraped into cold PBS. Cells were then pelleted and washed several times with cold PBS prior to processing [71]. Briefly, cells were lysed by sonication, dialysed and lyophilised. Reduction and carboxymethylation reactions were carried out on the lyophilised material, which was subsequently dialysed and lyophilised. Samples were digested with trypsin (Sigma) and the resultant glycopeptides purified by Sep-Pak C18 chromatography. N-glycans were released by PNGase-F (Roche) and purified by Sep-Pak C18 chromatography. N-glycan samples were then split into three aliquots. Two of the three aliquots were treated with either sialidase-S or sialidase-A (Agilent Technologies) and then purified by C18 Sep-Pak chromatography, the third aliquot was untreated. Untreated and treated glycans were then permethylated and analysed using a 4800 MALDI-TOF mass spectrometer (Applied Biosystems).

MS and MS/MS data were visualised using Data Explorer (Applied Biosystems) and [M + Na]+ molecular ions were annotated manually with the assistance of GlycoWorkbench [72]. Glycan structures were predicted based on composition and knowledge of biosynthetic pathways, with MS/MS analysis used to confirm assignments. Figures were prepared using Adobe Illustrator.

Glycomic data was quantified using an R programme described in Wu et al., 2024 [73] which detects glycans according to their isotopic peak pattern and calculates the relative intensity across the isotope cluster. An R programme was then used to calculate the proportions of a-2,3 and a-2,6 sialic acid. First, the relative intensity of each glycan was normalised to the sum of the relative intensities of all the glycans in the sample. For sialylated glycans, normalised intensities were then weighted based on the number of sialic acid residues present on each glycan. The proportion of a-2,6 linked sialic acid was determined by summing the weighted normalised intensities of sialylated molecular ion peaks remaining after sialidase-S digestion. The proportion of a-2,3 linked sialic acid was calculated by subtracting the weighted normalised intensities of sialylated molecular ions in the sialidase-S sample from the weighted normalised intensities of sialylated molecular ions from the untreated sample (raw data in Supplementary Table 2) [74].

Cell staining and flow cytometry

Infected bMECs were trypsinised, resuspended and fixed in PBS/4% formaldehyde. Following a PBS wash by centrifugation, cells were blocked with PBS/2%BSA for 1h. Cells were stained with rabbit anti-H3 (c33) or mouse anti-H1 (2-12C, Absolute antibody) primary antibodies diluted 1:5000 in blocking buffer for 1h at room temperature and with constant agitation. Cells were then washed twice 2x with blocking buffer and resuspended in a secondary antibody solution containing donkey anti-rabbit647 (ab150075, abcam) and donkey anti-mouse 488 (A11011, Invitrogen) for another 30 min. Cells were centrifuged, washed twice with blocking buffer and analysed using a BD LSR Fortessa X20 cytometer. The gating strategy used to define the infected cell populations is shown in Figure S13.